### 2023-02-02 data_processing_LML.qmd
Initial data load

---
title: "Data cleaning and processing code"
author: "Leah Lariscy"
format: html
editor: visual
---

## Load necessary packages here

```{r}
library(tidyverse)
library(skimr)
```

## Load the data here

```{r}
# Load CFX N1 data
n1 <- read_csv("data/raw_data/n1_data.csv")
# Load CFX N2 data
n2 <- read_csv("data/raw_data/n2_data.csv")
# Load Plant data
plant <- read_csv("data/raw_data/plant_data.csv")
# Load Standard curve data
std_curve <- read_csv("data/raw_data/standard_curves.csv")
# Load 
read_csv("data/raw_data/")
```

### 2023-02-03 code/processing_code/data_processing_LML.qmd
part 1 ready

---
title: "Data cleaning and processing script"
author: "Leah Lariscy"
format: html
editor: visual
---

## Data introduction

For this project, I would like to work with my own real-world data to explore the relationship between SARS-CoV-2 waste water surveillance data and COVID-19 clinical case data.

### Hypotheses so far

Waste water surveillance data will begin to correlate less with the clinical data as we progress further into the pandemic. (as rapid, at-home testing becomes more prevalent)

Catchment-level surveillance data will correlate better to waste water catchment-level case data than it will with county-level data.

## Load necessary packages here

```{r}
library(tidyverse)
library(skimr)
library(here)
```

## Load the data here

```{r}
here()
# Load CFX N1 data
n1 <- read_csv(here("data/raw_data/cfx_n1.csv"))

# Load CFX N2 data
n2 <- read_csv(here("data/raw_data/cfx_n2.csv"))

# Load Plant data
plant <- read_csv(here("data/raw_data/plant_data.csv"))

# Load COVID-19 Symptom data
covid_symptom <- read_csv(here("data/raw_data/ga_covid_data/epicurve_symptom_date.csv")) %>% 
  filter(county=="Clarke") %>% 
  select(symptom.date=`symptom date`, 
         cases, moving_avg_cases)

#Load COVID-19 Confirmed Case Data
covid_case <- read_csv(here("data/raw_data/ga_covid_data/epicurve_rpt_date.csv")) %>% 
  filter(county=="Clarke") %>% 
  select(report_date, 
         cases, 
         moving_avg_cases)

#Load COVID-19 Testing Data
covid_testing <- read_csv(here("data/raw_data/ga_covid_data/pcr_antigen_col.csv")) %>% 
  filter(county=="Clarke") %>% 
  select(collection_date = collection_dt, 
         pcr_tests = `ALL PCR tests performed`, 
         pcr_pos = `All PCR positive tests`, 
         pcr_pos_7dma = `7 day percent positive`,
         pcr_pos_14dma = `14 day percent positive`)

#Load Catchment-level COVID-19 Case Data
#covid_case_wrf = read_csv(here("data/raw_data/cases_wrf.csv"))
```

## Description of the data

### CFX N1 Data

`n1` \<- Waste water extracted RNA RT-qPCR results of SARS-CoV-2 N1 gene target

```{r}
skim(n1)
```

### CFX N2 Data

`n2` \<- Waste water extracted RNA RT-qPCR results of SARS-CoV-2 N2 gene target

```{r}
skim(n2)
```

### Plant Data

`plant` \<- Flow data from three Water Reclamation Facilities (WRF) in Athens-Clarke County (ACC).

```{r}
skim(plant)
```

### County-level COVID-19 Symptom Data

`covid_symptom` \<- Count of individuals who reported having symptoms of COVID-19 on a given day, subset to include only data from ACC. Accessed from Georgia Department of Public Health (GDPH) on 2023-01-31.

```{r}
skim(covid_symptom)
```

### County-level COVID-19 Case Data

`covid_case` \<- Count of confirmed cases of COVID-19 on a given day, subset to include only data from ACC. Accessed from GDPH on 2023-01-31.

```{r}
skim(covid_case)
```

### County-level COVID-19 Testing Data

`covid_testing` \<- Count of COVID-19 tests administered on a given day, subset to include only data from ACC. Accessed from GDPH on 2023-01-31.

```{r}
skim(covid_testing)
```

### Catchment-level COVID-19 Case Data

`covid_case_wrf` \<- Count of confirmed cases of COVID-19 on a given day from each WRF catchment zone in ACC.

```{r}
#I don't have access to this data yet, waiting on DPH partner to send.
```

### 2023-02-23 code/processing_code/data_processing_LML.qmd
data cleaning

@@
 editor: visual
+theme: solar
 ---
@@
 
-For this project, I would like to work with my own real-world data to explore the relationship between SARS-CoV-2 waste water surveillance data and COVID-19 clinical case data.
+For this project, I would like to work with my own real-world data to explore the relationship between **SARS-CoV-2 waste water surveillance data** and **COVID-19 clinical case data**.
 
@@
 
-Catchment-level surveillance data will correlate better to waste water catchment-level case data than it will with county-level data.
+Catchment-level surveillance data will correlate better to waste water catchment-level case data than it will with county-level data because people are more likely to poop at their own house and SARS-CoV-2 particles are shed more often in feces than they are in urine (Kim et al.
+2020).
 
@@
 ```
+
+## Data cleaning and wrangling
+
+Currently, I have 7 unique sets of data. In order to do analysis on the data down the road, I want to figure out how to best combine these data sets.
+
+Here is a list of the data sets I have:
+
+1.  CFX N1 results
+2.  CFX N2 results
+3.  Plant flow and total suspended solids
+4.  County-level COVID symptoms
+5.  County-level COVID confirmed cases
+6.  County-level COVID tests administered
+7.  Catchment-level COVID confirmed cases
+
+I will start by combining the CFX output data with the plant flow data and calling the dataframe `wbe`
+
+Next, I will combine all COVID-related data
+
+### Calculate SARS-CoV-2 concentrations from CFX data for both N1 and N2
+
+```{r}
+#N1 data here
+n1 <- n1 %>% mutate(ct = as.numeric(ct), 
+              copy_num_uL_rxn = as.numeric(10^((ct-36.046)/-3.5293)), #use standard curve slope here
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004)) #replace NAs with limit of detection (0.004 copies per uL of the reaction)
+
+#N2 data here (because they have different standard curve equations)
+n2 <- n2 %>% mutate(ct = as.numeric(ct),
+              copy_num_uL_rxn = as.numeric(10^((ct-37.731)/-3.2505)),
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))
+
+```
+
+### Combining and cleaning CFX data
+
+```{r}
+
+wbe_year2 <- bind_rows(n1, n2) %>% #combine dataframes with same number of rows
+  mutate(
+    date=as.Date(sample_date, format = "%d-%b-%y"), #change to date format
+    facility=substr(sample_id, 1,2), #first two letters in sample_id is facility ID
+    biological_replicate=substr(sample_id, nchar(sample_id), nchar(sample_id)), #last number in sample_id is the replicate number
+    ct=as.numeric(ifelse(ct=="Undetermined", NA, ct)) #set NAs
+  ) %>% 
+  arrange(date, facility, target, biological_replicate) %>% 
+  select(date, facility, target, biological_replicate, collection_num, ct, copy_num_uL_rxn) %>%
+  mutate(facility = as.factor(facility), facility = recode(facility, NO = "A", MI = "B", CC = "C"), facility = ordered(facility, levels = c("A", "B", "C"))) %>% #set facility to factor and then set factor levels
+  filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21"))) 
+```
+
+### Cleaning plant data
+
+```{r}
+plant_clean <- plant %>% 
+  mutate(date = as.Date(date, format = "%m/%d/%Y"), #change to date format
+         influent_flow_L = influent_flow_mg*1e6*231*(0.0254^3)*1000) %>%  #convert flow 
+  mutate(facility = as.factor(wrf), facility = recode(facility, NO = "A", MI = "B", CC = "C"), facility = ordered(facility, levels = c("A", "B", "C"))) %>% #set facility to factor and then set factor levels
+   select(date, facility, influent_flow_L, influent_tss_mg_l)
+
+```
+
+### Combining CFX and plant data
+
+```{r}
+n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date","facility"))
+```
+
+### Combining all COVID data
+
+```{r}
+covid <- full_join(
+  covid%>%
+    select(cases.symptom.onset=cases, date=symptom.date), 
+  covid.report%>%
+    select(cases.reported=cases, date=report_date), 
+  by = "date"
+) %>% 
+  full_join(
+    covid.testing%>%
+      rename(date=collection_date), 
+    by="date"
+  ) %>%
+  select(date, cases.symptom.onset, cases.reported, pcr_tests, pcr_pos, pcr_pos_7dma, pcr_pos_14dma) 
+  #full_join(cases.wrf, by = "date")
+```
+
+### Combining `n1_n2_plant` and `covid`
+
+```{r}
+all_data <- full_join(n1_n2_plant, covid, by = "date")
+```

### 2023-02-23 code/analysis_code/exploratory_analysis_LML.qmd
exploratory analysis

---
title: "Exploratory analysis of year 2 surveillance data"
author: "Leah Lariscy"
format: html
editor: visual
theme: solar
---

## Load necessary packages

```{r}
library(tidyverse)
library(here)
library(ggpmisc)

```

## Load data frames generated in data processing script

```{r}
here()
n1_n2_clean <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
plant_clean <- readRDS(here("data/processed_data/year2_plant_clean.rds"))
n1_n2_plant_combine <- readRDS(here("data/processed_data/year2_cfx_plant_combined.rds"))
covid_combine <- readRDS(here("data/processed_data/all_covid_combined.rds"))
all_data_combine <- readRDS(here("data/processed_data/year2_all_data_combined.rds"))
recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))

```

### Look at `all_data_combine` and determine visualization to make

```{r}
tibble(all_data_combine)
```

## Data Viz

Some plots I want to generate initially are

1.  Case averages over time
2.  Total copies over time (N1 and N2)

### Look at case average over time

```{r}
all_data_combine %>% group_by(date) %>% 
  summarize(mean_7dma = mean(pcr_pos_7dma)) %>% 
  ggplot(aes(x=date, y=mean_7dma)) +
  geom_point() +
 stat_smooth(method = "gam")

```

### Look at log10 total copies in wastewater over time for N1

First, average total copies of biological replicates. Then, add total copies from each facility.

```{r}
all_data_combine %>% filter(target == "N1") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_total_copies = mean(total_copies)) %>% 
  ungroup(facility) %>% 
  summarize(sum_total_copies = sum(mean_total_copies)) %>% 
  ggplot(aes(x = date,y = log10(sum_total_copies))) +
  geom_point(color = "red") +
  stat_smooth(method = "gam")

```

### Look at log10 total copies in wastewater over time for N2

```{r}
all_data_combine %>% filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_total_copies = mean(total_copies)) %>% 
  ungroup(facility) %>% 
  summarize(sum_total_copies = sum(mean_total_copies)) %>% 
  ggplot(aes(date, log10(sum_total_copies))) +
  geom_point() +
  stat_smooth(method = "gam")
```

### Log10 total copies per day for N1 and N2

```{r}
all_data_combine %>% 
  drop_na() %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_total_copies = mean(total_copies)) %>% 
  ungroup(facility) %>% 
  summarize(sum_total_copies = sum(mean_total_copies)) %>% 
  ggplot(aes(date, log10(sum_total_copies), color = target)) +
  geom_point() +
  stat_smooth(method = "gam")
```

### Look at log10 copies per L in N1

```{r}
all_data_combine %>% filter(target == "N1") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
  ggplot(aes(date, log10(sum_mean_copies_L))) +
  geom_point() +
  stat_smooth(method = "gam")
```

### Look at log10 copies per liter in N1

```{r}
all_data_combine %>% filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
  ggplot(aes(date, log10(sum_mean_copies_L))) +
  geom_point() +
  stat_smooth(method = "gam")
```

### Look at log10 copies per liter across both N1 and N2

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
  ggplot(aes(date, log10(sum_mean_copies_L))) +
 stat_smooth(method = "gam")
```

### Look at log10 copies per liter vs case 7dma

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies_L), mean_7dma)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq()
```

### Look at log10 copies of N1 per liter vs case 7dma

```{r}
all_data_combine %>% filter(target == "N1") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies_L), mean_7dma)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq()
```

### Look at log10 copies of N2 per liter vs case 7dma

```{r}
all_data_combine %>% filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies_L), mean_7dma)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq()
```

### Look at relationship of copies/L and 7dma over time

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(copy_num_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(x=date)) +
  #geom_point(aes(y=log10(sum_mean_copies_L)/mean_7dma)) + 
  geom_line(aes(y=(log10(sum_mean_copies_L)/mean_7dma), color = "red")) +
  geom_line(aes(y=log10(mean_7dma)), color = "blue")

```

### 2023-02-23 code/processing_code/data_processing_LML.qmd
exploratory analysis

@@
 ## Load the data here
+
+recovery_output \<- read_csv(here("data/raw_data/recovery_data.csv"))
+
+recovery_input \<- read_csv(here("data/raw_data/calfguard.csv"))
 
@@
 #covid_case_wrf = read_csv(here("data/raw_data/cases_wrf.csv"))
+
+#Load CFX recovery data
+recovery_output <- read_csv(here("data/raw_data/recovery_data.csv"))
+recovery_input <- read_csv(here("data/raw_data/calfguard.csv"))
 ```
@@
 7.  Catchment-level COVID confirmed cases
+8.  CFX recovery input
+9.  CFX recovery output
 
@@
   select(date, facility, target, biological_replicate, collection_num, ct, copy_num_uL_rxn) %>%
-  mutate(facility = as.factor(facility), facility = recode(facility, NO = "A", MI = "B", CC = "C"), facility = ordered(facility, levels = c("A", "B", "C"))) %>% #set facility to factor and then set factor levels
+  mutate(facility = as.factor(facility), 
+         facility = recode(facility, NO = "A", MI = "B", CC = "C"), 
+         facility = ordered(facility, levels = c("A", "B", "C")), #set facility to factor and then set factor levels
+         copy_num_L = copy_num_uL_rxn*20/5*60/280*1000*1000) %>% 
   filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+
+saveRDS(wbe_year2, here("data/processed_data/year2_cfx_clean.rds"))
 ```
@@
 
+saveRDS(plant_clean, here("data/processed_data/year2_plant_clean.rds"))
 ```
@@
 ```{r}
-n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date","facility"))
+n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date","facility")) %>% 
+  mutate(total_copies = copy_num_L * influent_flow_L)
+
+saveRDS(n1_n2_plant, here("data/processed_data/year2_cfx_plant_combined.rds"))
 ```
@@
 covid <- full_join(
-  covid%>%
+  covid_symptom%>%
     select(cases.symptom.onset=cases, date=symptom.date), 
-  covid.report%>%
+  covid_case%>%
     select(cases.reported=cases, date=report_date), 
@@
   full_join(
-    covid.testing%>%
+    covid_testing%>%
       rename(date=collection_date), 
@@
   #full_join(cases.wrf, by = "date")
+
+saveRDS(covid, here("data/processed_data/all_covid_combined.rds"))
 ```
@@
 ```{r}
-all_data <- full_join(n1_n2_plant, covid, by = "date")
+all_data <- full_join(n1_n2_plant, covid, by = "date") %>% 
+   filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+
+saveRDS(all_data, here("data/processed_data/year2_all_data_combined.rds"))
 ```
+
+### Cleaning recovery data sets
+
+```{r}
+#Start with recovery_output 
+recovery_output_clean <- recovery_output %>% 
+  mutate(ct = as.numeric(ct)) %>% 
+  group_by(sample_id, cg_num) %>% 
+  summarize(avg_ct = mean(ct)) %>% 
+  separate(sample_id, into = c("facility", "collection_num", "rep"), sep = "_") %>% 
+  mutate(collection_num = as.numeric(collection_num),
+         copies_ul_rxn = 10^((avg_ct-30.7)/-3.238),
+         copies_ul_sample = copies_ul_rxn *20/2*25/3*60/280) %>% 
+  select(facility, collection_num, cg_num, copies_ul_sample, avg_ct) %>% 
+  drop_na()
+
+recovery_input_clean <- recovery_input %>% 
+  mutate(ct = as.numeric(ct)) %>% 
+  group_by(sample_id) %>% 
+  summarize(avg_ct = mean(ct)) %>% 
+  separate(sample_id, into = c("CG", "cg_num"), sep = "_") %>% 
+  mutate(cg_num = as.numeric(cg_num),
+         copies_ul_rxn = 10^((avg_ct-30.7)/-3.238),
+         copies_ul_input = copies_ul_rxn *20/2*25/3*60/50*40/40000) %>% 
+  select(cg_num, copies_ul_input, avg_ct) %>% 
+  drop_na()
+
+saveRDS(recovery_output_clean, here("data/processed_data/recovery_output_clean.rds"))
+saveRDS(recovery_input_clean, here("data/processed_data/recovery_input_clean.rds"))
+
+```
+
+### Combining recovery data sets
+
+```{r}
+recovery_calc <- left_join(recovery_output_clean, recovery_input_clean, by = "cg_num") %>% 
+  mutate(percent_recovery = (copies_ul_sample/copies_ul_input)*100) %>% 
+  select(facility, collection_num, percent_recovery)
+
+saveRDS(recovery_calc, here("data/processed_data/recovery_percentages.rds"))
+```

### 2023-02-24 code/analysis_code/exploratory_analysis_LML.qmd
exploratory analysis

@@
 recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))
+n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))
 
@@
 
-### Look at `all_data_combine` and determine visualization to make
+### Look at `all_data_combine`
 
@@
 
-Some plots I want to generate initially are
+### Variation analysis
 
-1.  Case averages over time
-2.  Total copies over time (N1 and N2)
+Graphs to make:
 
-### Look at case average over time
+-   histogram of case count observations
+
+-   histogram of 7 day moving average case observations
+
+-   histogram of total daily copies
+
+-   histogram of copies per L
+
+-   histogram of percent positive qPCR assays
+
+### Co-variation analysis
+
+Graphs to make:
+
+#### Histogram of clinical case 7 day moving average
+
+```{r}
+covid_combine %>% ggplot(aes(pcr_pos_7dma)) +
+  geom_histogram()
+summary(covid_combine$pcr_pos_7dma)
+```
+
+#### Histogram of PCR positive clinical cases
+
+```{r}
+covid_combine %>% ggplot(aes(pcr_pos)) +
+  geom_histogram()
+summary(covid_combine$pcr_pos)
+```
+
+#### Histogram of clinical case 14 day moving average
+
+```{r}
+covid_combine %>% ggplot(aes(pcr_pos_14dma)) +
+  geom_histogram()
+summary(covid_combine$pcr_pos_14dma)
+```
+
+#### Histogram of total daily viral copies at county level
+
+```{r}
+sum_daily_copies_county <- all_data_combine %>% 
+  drop_na() %>% 
+  group_by(date, facility) %>% 
+  summarize(mean_total_copies = mean(total_copies)) %>% 
+  ungroup(facility) %>% 
+  summarize(sum_total_copies = sum(mean_total_copies))
+
+  sum_daily_copies_county %>% ggplot(aes(log10(sum_total_copies))) +
+  geom_histogram()
+
+  log10_sum_daily_copies <- log10(sum_daily_copies_county$sum_total_copies)
+  summary(log10_sum_daily_copies)
+```
+
+### Look at case 7 day moving average over time
 
@@
 
-### Look at log10 total copies in wastewater over time for N1
-
-First, average total copies of biological replicates. Then, add total copies from each facility.
+Case 7 day moving average overtime, bar plot and GAM model
 
 ```{r}
-all_data_combine %>% filter(target == "N1") %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_total_copies = mean(total_copies)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_total_copies = sum(mean_total_copies)) %>% 
-  ggplot(aes(x = date,y = log10(sum_total_copies))) +
-  geom_point(color = "red") +
-  stat_smooth(method = "gam")
+all_data_combine %>% group_by(date) %>% 
+  summarize(mean_7dma = mean(pcr_pos_7dma),
+            mean_pcr_pos = mean(pcr_pos)) %>% 
+  ggplot(aes(x = date, y = mean_7dma)) +
+  geom_bar(stat = "identity", color = '#634490') +
+  stat_smooth(method = "gam", color = 'black')
 
@@
 
-### Look at log10 total copies in wastewater over time for N2
+PCR positives per day over time, bar plot and GAM model
 
 ```{r}
-all_data_combine %>% filter(target == "N2") %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_total_copies = mean(total_copies)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_total_copies = sum(mean_total_copies)) %>% 
+all_data_combine %>% group_by(date) %>% 
+  summarize(mean_pcr_pos = mean(pcr_pos)) %>% 
+  ggplot(aes(x = date, y = mean_pcr_pos)) +
+  geom_bar(stat = "identity", color = '#634490') +
+  stat_smooth(method = "gam", color = 'black') 
+```
+
+### Look at log10 total copies in wastewater over time (avg N1/N2)
+
+```{r}
+all_data_combine %>% #filter(target == "N2") %>% 
+  #drop_na() %>% #remove NAs from analysis
+  group_by(date, facility) %>% #this will allow me to average the two targets
+  summarize(mean_total_copies = mean(total_copies)) %>% #average across N1 and N2
+  ungroup(facility) %>% #need to do this so I can add the avg total copies of each WRF
+  summarize(sum_total_copies = sum(mean_total_copies)) %>% #total daily county copies
   ggplot(aes(date, log10(sum_total_copies))) +
@@
 all_data_combine %>% 
-  drop_na() %>% 
   group_by(date, target, facility) %>% 
+  drop_na(target) %>% 
   summarize(mean_total_copies = mean(total_copies)) %>% 
@@
   summarize(sum_total_copies = sum(mean_total_copies)) %>%
+
   ggplot(aes(date, log10(sum_total_copies), color = target)) +
@@
 all_data_combine %>% filter(target == "N1") %>%
-  drop_na() %>% 
+
   group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(copy_num_L)) %>% 
+  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
   ungroup(facility) %>% 
   summarize(sum_mean_copies_L = sum(mean_copies_L)) %>%
+
   ggplot(aes(date, log10(sum_mean_copies_L))) +
@@
 
-### Look at log10 copies per liter in N1
+### Look at log10 copies per liter in N2
 

### 2023-02-24 code/processing_code/data_processing_LML.qmd
exploratory analysis

@@
 
-### Hypotheses so far
+## Questions I have so far
 
-Waste water surveillance data will begin to correlate less with the clinical data as we progress further into the pandemic. (as rapid, at-home testing becomes more prevalent)
+### Variation
 
-Catchment-level surveillance data will correlate better to waste water catchment-level case data than it will with county-level data because people are more likely to poop at their own house and SARS-CoV-2 particles are shed more often in feces than they are in urine (Kim et al.
-2020).
+1.  What is the distribution like of SARS-CoV-2 viral load observations? What does this distribution look like at catchment levels vs at the county level?
+2.  What is the distribution like of N1 and N2 qPCR percent positive observations? What does this distribution look like at catchment levels vs at the county level?
+3.  What is the distribution like of clinical case observations? What does this distribution look like at catchment levels vs at the county level?
+
+### Covariation
+
+1.  Is there a relationship between viral load and clinical case observations?
+
+2.  Is there a relationship between qPCR percent positive and clinical case observations?
+
+3.  What do viral load trends look like over time?
+
+4.  What do clinical case trends look like over time?
+
+5.  How do each of these temporal trends compare to one another over time?
 
@@
 
-I will start by combining the CFX output data with the plant flow data and calling the dataframe `wbe`
+I will start by combining the CFX output data with the plant flow data and calling the data frame `wbe`
 
@@
 
+### Summarizing CFX data for downstream processing
+
+```{r}
+wbe_year2_avg <- wbe_year2 %>% group_by(date,facility,target,collection_num) %>% 
+  summarize(avg_copy_L = mean(copy_num_L), 
+            avg_copy_uL_rxn = mean(copy_num_uL_rxn)) %>% 
+  ungroup()
+saveRDS(wbe_year2_avg, here("data/processed_data/year2_cfx_avg.rds"))
+```
+
 ### Cleaning plant data
@@
 ```{r}
-n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date","facility")) %>% 
-  mutate(total_copies = copy_num_L * influent_flow_L)
+n1_n2_plant <- left_join(wbe_year2_avg, plant_clean, by = c("date","facility")) %>% 
+  mutate(total_copies = avg_copy_L * influent_flow_L)
 

### 2023-02-27 code/analysis_code/exploratory_analysis_LML.qmd
exploratory analysis

@@
 library(ggpmisc)
+library(ggthemes)
 
@@
 covid_combine %>% ggplot(aes(pcr_pos_7dma)) +
-  geom_histogram()
+  geom_histogram(fill = "#C06C84") +
+xlab(label = "Clinical Case Seven Day Moving Average") +
+ylab(label = "Count") +
+theme_light()
+
+ggsave(here("figures/exploratory/variate/7DMA_histo.png"))
 summary(covid_combine$pcr_pos_7dma)
@@
 covid_combine %>% ggplot(aes(pcr_pos)) +
-  geom_histogram()
+  geom_histogram(fill = '#F8B195') +
+  xlab(label = "Daily Positive Clinical Cases") +
+  ylab(label = "Count") +
+  theme_light()
+
+ggsave(here("figures/exploratory/variate/pos_cases_histo.png"))
 summary(covid_combine$pcr_pos)
@@
   drop_na() %>% 
-  group_by(date, facility) %>% 
+  group_by(date, target, facility) %>% 
   summarize(mean_total_copies = mean(total_copies)) %>% 
@@
 
-  sum_daily_copies_county %>% ggplot(aes(log10(sum_total_copies))) +
-  geom_histogram()
+  sum_daily_copies_county %>% ggplot(aes(log10(sum_total_copies), fill = target)) +
+  geom_histogram() +
+      scale_fill_manual(name = "Target",values = c("#F67280", "#355C7D")) +
+    xlab(label = "Log10 Total Daily County-Level Copies") +
+    ylab(label = "Count") +
+    theme_light()
 
+ggsave(here("figures/exploratory/variate/total_copies_histo.png"))
   log10_sum_daily_copies <- log10(sum_daily_copies_county$sum_total_copies)
   summary(log10_sum_daily_copies)
-```
-
-### Look at case 7 day moving average over time
-
-```{r}
-all_data_combine %>% group_by(date) %>% 
-  summarize(mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ggplot(aes(x=date, y=mean_7dma)) +
-  geom_point() +
- stat_smooth(method = "gam")
-
 ```
@@
   ggplot(aes(x = date, y = mean_7dma)) +
-  geom_bar(stat = "identity", color = '#634490') +
-  stat_smooth(method = "gam", color = 'black')
+  geom_bar(stat = "identity", color = '#C06C84') +
+  stat_smooth(method = "gam", color = 'black') +
+  xlab(label = "Date") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/7DMA_time_plot.png"))
 
@@
   ggplot(aes(x = date, y = mean_pcr_pos)) +
-  geom_bar(stat = "identity", color = '#634490') +
-  stat_smooth(method = "gam", color = 'black') 
+  geom_bar(stat = "identity", color = '#F8B195') +
+  stat_smooth(method = "gam", color = 'black') +
+  ylab(label = "Positive Cases Per Day") +
+  xlab(label = "Date") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/pos_cases_time.png"))
 ```
@@
 all_data_combine %>% #filter(target == "N2") %>% 
-  #drop_na() %>% #remove NAs from analysis
+  drop_na(target) %>% #remove NAs from analysis
   group_by(date, facility) %>% #this will allow me to average the two targets
@@
   ggplot(aes(date, log10(sum_total_copies))) +
-  geom_point() +
-  stat_smooth(method = "gam")
+  geom_point(color = "#80256D") +
+  stat_smooth(method = "gam", color = "#80256D") +
+   xlab(label = "Date") +
+  ylab(label = "Log10 Total Daily County-Level Copies") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/n1_n2_avg_time.png"))
 ```
@@
   geom_point() +
-  stat_smooth(method = "gam")
+  stat_smooth(method = "gam") +
+  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
+  xlab(label = "Date") +
+  ylab(label = "Log10 Total Daily County-Level Copies") +
+  theme_light()
+
+ ggsave(here("figures/exploratory/covariate/n1_n2_time.png"))
 ```
@@
   group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(copy_num_L)) %>% 
+  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
   ungroup(facility) %>% 
@@
   drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(copy_num_L)) %>% 
+  group_by(date, target, facility) %>% 
+  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
   ungroup(facility) %>% 
   summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
-  ggplot(aes(date, log10(sum_mean_copies_L))) +
- stat_smooth(method = "gam")
+  ggplot(aes(date, log10(sum_mean_copies_L),color = target)) +
+ stat_smooth(method = "gam") +
+  geom_point()
+
 ```
@@
   group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(copy_num_L), 
+  summarize(mean_copies_L = mean(avg_copy_L), 
             mean_7dma = mean(pcr_pos_7dma)) %>% 
@@
 
+### Look at log10 total daily copies per 7DMA
+
+```{r}
+all_data_combine %>% #filter(target == "N2") %>% 
+  drop_na() %>% 
+  group_by(date, facility) %>% 
+  summarize(mean_daily_copies = mean(total_copies), 
+            mean_7dma = mean(pcr_pos_7dma)) %>% 
+  ungroup(facility) %>% 
+  summarize(sum_mean_copies = sum(mean_daily_copies),
+            mean_7dma = mean(mean_7dma)) %>% 
+  ggplot(aes(log10(sum_mean_copies), mean_7dma)) +
+  geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
+  stat_poly_eq() +
+  xlab(label = "Log10 Total Daily County-Level Copies") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/total_copies_avg_vs_7DMA.png"))
+```
+
+```{r}
+all_data_combine %>% 
+  drop_na() %>% 
+  group_by(date, target, facility) %>% 
+  summarize(mean_daily_copies = mean(total_copies), 
+            mean_7dma = mean(pcr_pos_7dma)) %>% 
+  ungroup(facility) %>% 
+  summarize(sum_mean_copies = sum(mean_daily_copies),
+            mean_7dma = mean(mean_7dma)) %>% 
+  ggplot(aes(log10(sum_mean_copies), mean_7dma, color = target)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
+  xlab(label = "Log10 Total Daily County-Level Copies") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/total_copies_vs_7DMA.png"))
+```
+
 ### Look at relationship of copies/L and 7dma over time
@@
 ```
+
+### Log10 Daily copies N1 & N2 vs daily positive cases
+
+```{r}
+all_data_combine %>% 
+  drop_na() %>% 
+  group_by(date, target, facility) %>% 
+  summarize(mean_daily_copies = mean(total_copies), 
+            mean_pcr_pos = mean(pcr_pos)) %>% 
+  ungroup(facility) %>% 
+  summarize(sum_mean_copies = sum(mean_daily_copies),
+            mean_pcr_pos = mean_pcr_pos) %>% 
+  ggplot(aes(log10(sum_mean_copies), mean_pcr_pos, color = target)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
+  xlab(label = "Log10 Total Daily County-Level Copies") +
+  ylab(label = "Daily Positive Clinical Cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/total_copies_vs_pcr_pos.png"))
+```
+
+### Log10 Daily copies average vs daily positive cases
+
+```{r}
+all_data_combine %>% 
+  drop_na() %>% 
+  group_by(date, facility) %>% 
+  summarize(mean_daily_copies = mean(total_copies), 
+            mean_pcr_pos = mean(pcr_pos)) %>% 
+  ungroup(facility) %>% 
+  summarize(sum_mean_copies = sum(mean_daily_copies),
+            mean_pcr_pos = mean_pcr_pos) %>% 
+  ggplot(aes(log10(sum_mean_copies), mean_pcr_pos)) +
+  geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
+  stat_poly_eq() +
+  xlab(label = "Log10 Total Daily County-Level Copies") +
+  ylab(label = "Daily Positive Clinical Cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/total_copies_avg_vs_pcr_pos.png"))
+```

### 2023-03-02 code/processing_code/data_processing_LML.qmd
exploratory analysis

@@
 ```
+
+### Calculate percent positive of technical reps
+
+```{r}
+wbe_tr <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
+  dplyr::summarise(
+    n=n(), 
+    n.miss=sum(is.na(ct)), 
+    ct.mean=mean(ct,na.rm=T), 
+    ct.sd=sd(ct,na.rm=T)
+  ) %>% 
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(n.pos = n - n.miss) %>% 
+  mutate(pos = ifelse(n.pos > 0, "pos", "neg"))
+wbe_tr
+saveRDS(wbe_tr,here("data/processed_data/wbe_technical_reps.rds"))
+```
+
+### Calculate percent positive per biological replicate
+
+```{r}
+wbe_br <- wbe_tr %>%
+  group_by(date, facility, target) %>% 
+  dplyr::summarise(
+    n.bio = n(),
+    n.bio.non.miss = sum(!is.na(ct.mean)),
+    n.bio.miss = sum(is.na(ct.mean)),
+    n.total = sum(n), 
+    n.total.miss = sum(n.miss), 
+    bio.ct.mean = mean(ct.mean, na.rm = T), 
+    bio.ct.sd = sd(ct.mean, na.rm=T), 
+    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
+  ) %>%
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup()
+wbe_br
+```
+
+### Calculate percent positive per sample
+
+```{r}
+wbe_samp <- wbe_br %>%
+  group_by(date, facility, target) %>% 
+  dplyr::summarise(
+    n.pos = sum(!is.na(bio.ct.mean)),
+    n.miss = sum(is.na(bio.ct.mean))) %>%
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(prop.pos = n.pos/(n.pos+n.miss)*100)
+wbe_samp
+```
+
+```{r}
+ggstatsplot::ggbarstats(
+  data = wbe_samp,
+  x = n.pos,
+  y = target) +
+  labs(caption = NULL)
+```

### 2023-03-13 docs/code/analysis_code/exploratory_analysis_LML.qmd
removed unnecessary files

---
title: "Exploratory analysis of year 2 surveillance data"
author: "Leah Lariscy"
format: html
editor: visual
theme: solar
---

## Load necessary packages

```{r}
library(tidyverse)
library(here)
library(ggpmisc)
library(ggthemes)

```

## Load data frames generated in data processing script

```{r}
here()
n1_n2_clean <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
plant_clean <- readRDS(here("data/processed_data/year2_plant_clean.rds"))
n1_n2_plant_combine <- readRDS(here("data/processed_data/year2_cfx_plant_combined.rds"))
covid_combine <- readRDS(here("data/processed_data/all_covid_combined.rds"))
all_data_combine <- readRDS(here("data/processed_data/year2_all_data_combined.rds"))
recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))
n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))

```

### Look at `all_data_combine`

```{r}
tibble(all_data_combine)
```

# Data Viz

### Variation analysis

Graphs to make:

-   histogram of case count observations

-   histogram of 7 day moving average case observations

-   histogram of total daily copies

-   histogram of copies per L

-   histogram of percent positive qPCR assays

### Co-variation analysis

Graphs to make:

## Single variate analysis

### Clinical case 7 day moving average histogram

```{r}
covid_combine %>% ggplot(aes(pcr_pos_7dma)) +
  geom_histogram(fill = "#C06C84") +
xlab(label = "Clinical Case Seven Day Moving Average") +
ylab(label = "Count") +
theme_light()

ggsave(here("figures/exploratory/variate/7DMA_histo.png"))
summary(covid_combine$pcr_pos_7dma)
```

### PCR positive clinical cases histogram

```{r}
covid_combine %>% ggplot(aes(pcr_pos)) +
  geom_histogram(fill = '#F8B195') +
  xlab(label = "Daily Positive Clinical Cases") +
  ylab(label = "Count") +
  theme_light()

ggsave(here("figures/exploratory/variate/pos_cases_histo.png"))
summary(covid_combine$pcr_pos)
```

### Clinical case 14 day moving average histogram

```{r}
covid_combine %>% ggplot(aes(pcr_pos_14dma)) +
  geom_histogram(fill = "#FF7E5A")

summary(covid_combine$pcr_pos_14dma)

ggsave(here("figures/exploratory/variate/14DMA_histo.png"))

```

### Total copies in WW at county level histogram

```{r}
sum_daily_copies_county <- all_data_combine %>% 
  drop_na() %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_total_copies = mean(total_copies)) %>% 
  ungroup(facility) %>% 
  summarize(sum_total_copies = sum(mean_total_copies))

  sum_daily_copies_county %>% ggplot(aes(log10(sum_total_copies), fill = target)) +
  geom_histogram() +
      scale_fill_manual(name = "Target",values = c("#F67280", "#355C7D")) +
    xlab(label = "Log10 Total Daily County-Level Copies") +
    ylab(label = "Count") +
    theme_light()

ggsave(here("figures/exploratory/variate/total_copies_histo.png"))
  log10_sum_daily_copies <- log10(sum_daily_copies_county$sum_total_copies)
  summary(log10_sum_daily_copies)
```

### Copies per liter histogram

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>%

  ggplot(aes(log10(sum_mean_copies_L), fill = target)) +
  geom_histogram() +
      scale_fill_manual(name = "Target",values = c("#F67280", "#355C7D")) +
    xlab(label = "Log10 Copies Per Liter") +
    ylab(label = "Count") +
    theme_light()

ggsave(here("figures/exploratory/variate/copies_L_histo.png"))

```

## Co-variate analysis: time series

### Case 7 day moving average overtime, bar plot and GAM model

```{r}
all_data_combine %>% group_by(date) %>% 
  summarize(mean_7dma = mean(pcr_pos_7dma),
            mean_pcr_pos = mean(pcr_pos)) %>% 
  ggplot(aes(x = date, y = mean_7dma)) +
  geom_point(color = '#C06C84') +
  geom_line(color = '#C06C84') +
  xlab(label = "Date") +
  ylab(label = "Clinical Case 7-Day Moving Average") +
  theme_light()

ggsave(here("figures/exploratory/covariate/7DMA_time_plot.png"))

```

### PCR positives per day over time, bar plot and GAM model

```{r}
all_data_combine %>% group_by(date) %>% 
  summarize(mean_pcr_pos = mean(pcr_pos)) %>% 
  ggplot(aes(x = date, y = mean_pcr_pos)) +
  geom_bar(stat = "identity", color = '#F8B195') +
  geom_line(color = 'black') +
  ylab(label = "Positive Cases Per Day") +
  xlab(label = "Date") +
  theme_light()

ggsave(here("figures/exploratory/covariate/pos_cases_time.png"))
```

### Case 14 day moving average over time

```{r}
all_data_combine %>% group_by(date) %>% 
  summarize(mean_14_dma = mean(pcr_pos_14dma)) %>%

   ggplot(aes(x = date, y = mean_14_dma)) +
  geom_bar(stat = "identity", color = '#FF7E5A') +
  stat_smooth(method = "gam", color = 'black') +
  ylab(label = "Clinical Case 14-Day Moving Average") +
  xlab(label = "Date") +
  theme_light()

ggsave(here("figures/exploratory/covariate/14DMA_time.png"))
```

### Log10 total copies in wastewater over time (avg N1/N2)

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na(target) %>% #remove NAs from analysis
  group_by(date, facility) %>% #this will allow me to average the two targets
  summarize(mean_total_copies = mean(total_copies)) %>% #average across N1 and N2
  ungroup(facility) %>% #need to do this so I can add the avg total copies of each WRF
  summarize(sum_total_copies = sum(mean_total_copies)) %>% #total daily county copies
  ggplot(aes(date, log10(sum_total_copies))) +
  geom_point(color = "#80256D") +
  geom_line(color = "#80256D") +
   xlab(label = "Date") +
  ylab(label = "Log10 Total Daily County-Level Copies") +
  theme_light()

ggsave(here("figures/exploratory/covariate/avg_total_copies_time.png"))
```

### Log10 viral load in wastewater for N1 and N2

```{r}
all_data_combine %>% 
  group_by(date, target, facility) %>% 
  drop_na(target) %>% 
  summarize(mean_total_copies = mean(total_copies)) %>% 
  ungroup(facility) %>% 
  summarize(sum_total_copies = sum(mean_total_copies)) %>%

  ggplot(aes(date, log10(sum_total_copies), color = target)) +
  geom_point() +
  geom_line() +
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Date") +
  ylab(label = "Log10 County-Level Viral Load") +
  theme_light() +
  facet_wrap(vars(target),nrow = 2)

 ggsave(here("figures/exploratory/covariate/total_copies_time.png"))
```

### Log10 copies per liter across time (both N1 and N2)

```{r}
all_data_combine %>%
  drop_na(target) %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
  ggplot(aes(date, log10(sum_mean_copies_L),color = target)) +
geom_line() +
  geom_point() + 
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Date") +
  ylab(label = "Log10 County-Level Copies Per Liter") +
  theme_light() +
    facet_wrap(vars(target),nrow = 2)

 ggsave(here("figures/exploratory/covariate/copies_L_time.png"))

```

## Co-variate analysis: correlations between WW and clinical variables

### Log10 copies per liter vs case 7DMA (N1/N2 average)

```{r}
all_data_combine %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies_L), log10(mean_7dma))) +
  geom_point(color = "#80256D") +
  stat_poly_line(color = "#80256D") +
  stat_poly_eq(color = "#80256D") +
   xlab(label = "Log10 County-Level Copies Per Liter") +
  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
  theme_light()

ggsave(here("figures/exploratory/covariate/copies_L_avg_vs_7DMA.png"))
```

### Log10 copies per liter vs case 7DMA (N1 & N2)

```{r}
all_data_combine %>%  
  drop_na(target) %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies_L), log10(mean_7dma), color = target)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq() +
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Log10 County-Level Copies Per Liter") +
  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
  theme_light()

ggsave(here("figures/exploratory/covariate/copies_L_vs_7DMA.png"))
```

### Log10 Copies per Liter vs daily clinical cases

```{r}
all_data_combine %>%  
  drop_na(target) %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_pcr_pos = mean(pcr_pos)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_pcr_pos = mean_pcr_pos) %>% 
  ggplot(aes(log10(sum_mean_copies_L), log10(mean_pcr_pos), color = target)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq() +
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Log10 County-Level Copies Per Liter") +
  ylab(label = "Log10 Daily Positive Clinical Cases") +
  theme_light()

ggsave(here("figures/exploratory/covariate/copies_L_vs_pcr_pos.png"))
```

### Log10 Copies per Liter vs daily clinical cases (N1 N2 average)

```{r}
all_data_combine %>%  
  drop_na(target) %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_pcr_pos = mean(pcr_pos)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_pcr_pos = mean_pcr_pos) %>% 
  ggplot(aes(log10(sum_mean_copies_L), log10(mean_pcr_pos))) +
  geom_point(color = "#80256D") +
  stat_poly_line(color = "#80256D") +
  stat_poly_eq(color = "#80256D") +
  xlab(label = "Log10 County-Level Copies Per Liter") +
  ylab(label = "Log10 Daily Positive Clinical Cases") +
  theme_light()

ggsave(here("figures/exploratory/covariate/copies_L_avg_vs_pcr_pos.png"))
```

### Log10 viral load vs 7DMA

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_daily_copies = mean(total_copies), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies = sum(mean_daily_copies),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies), log10(mean_7dma))) +
  geom_point(color = "#80256D") +
  stat_poly_line(color = "#80256D") +
  stat_poly_eq(color = "#80256D") +
  xlab(label = "Log10 Daily Viral Load") +
  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
  theme_light()

ggsave(here("figures/exploratory/covariate/avg_total_copies_vs_7DMA.png"))
```

### Log10 viral load vs 7DMA (N1 & N2)

```{r}
all_data_combine %>% 
  drop_na() %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_daily_copies = mean(total_copies), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies = sum(mean_daily_copies),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(log10(sum_mean_copies), log10(mean_7dma), color = target)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq() +
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Log10 Daily Viral Load") +
  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
  theme_light()

ggsave(here("figures/exploratory/covariate/total_copies_vs_7DMA.png"))
```

### Look at relationship of copies/L and 7dma over time

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(x=date)) +
  #geom_point(aes(y=log10(sum_mean_copies_L)/mean_7dma)) + 
  geom_line(aes(y=(log10(sum_mean_copies_L)/mean_7dma), color = "red")) +
  geom_line(aes(y=log10(mean_7dma)), color = "blue")

```

### Log10 viral load vs daily positive cases

```{r}
all_data_combine %>% 
  drop_na() %>% 
  group_by(date, target, facility) %>% 
  summarize(mean_daily_copies = mean(total_copies), 
            mean_pcr_pos = mean(pcr_pos)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies = sum(mean_daily_copies),
            mean_pcr_pos = mean_pcr_pos) %>% 
  ggplot(aes(log10(sum_mean_copies), log10(mean_pcr_pos), color = target)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq() +
  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
  xlab(label = "Log10 Daily Viral Load") +
  ylab(label = "Log10 Daily Positive Clinical Cases") +
  theme_light()

ggsave(here("figures/exploratory/covariate/total_copies_vs_pcr_pos.png"))
```

### Log10 viral load vs daily positive cases

```{r}
all_data_combine %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_daily_copies = mean(total_copies), 
            mean_pcr_pos = mean(pcr_pos)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies = sum(mean_daily_copies),
            mean_pcr_pos = mean_pcr_pos) %>% 
  ggplot(aes(log10(sum_mean_copies), log10(mean_pcr_pos))) +
  geom_point(color = "#80256D") +
  stat_poly_line(color = "#80256D") +
  stat_poly_eq(color = "#80256D") +
  xlab(label = "Log10 Daily Viral Load") +
  ylab(label = "Log10 Daily Reported Clinical Cases") +
  theme_light()

ggsave(here("figures/exploratory/covariate/total_copies_avg_vs_pcr_pos.png"))
```

### 2023-03-13 products/manuscript/Manuscript.qmd
removed unnecessary files

---
title: "Example Manuscript Template for a Data Analysis Project"
subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
#library(here)
#library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
#knitr::include_graphics(here("results","height_weight_stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
#resulttable2 = readRDS(here("results","resulttable2.rds"))
#knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2023-03-13 products/manuscript/supplement/Supplementary_Material.qmd
removed unnecessary files

---
title: "Supplement to Example Manuscript Template for a Data Analysis Project"

subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../dataanalysis_template_references.bib
csl: ../../vancouver.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

This shows some materials that could go into a supplementary file. Often you want/need references here too. You can use the same reference bib file for this and the main text (as done here) or have separate bib files.

For illustrative purposes, I'm doing the supplement as pdf. For this to work, you need a (La)TeX system installed. It's easy. Just follow [these steps](https://quarto.org/docs/output-formats/pdf-basics.html).

Of course you would choose the format based on needs.

I'm also using a different style for the references here. (vancouver vs apa in the main manuscript). Usually one would have the formatting of the references the same in those two documents, but I want to illustrate how easy it is to switch reference formatting styles, you just need to get the right CSL file and specify it in the YAML header. We could also have a seperate reference bibtext (`.bib`) file, but here we are using the same.

# Overview

A quick overview of what readers can find in the supplement.

# Code and file information

Explain here what each code/file is and does, and in which order (if any) users need to run thing to reproduce everything.
Essentially, give a full set of instructions to re-generate everything.

{{< pagebreak >}}

# Additional Method Details

Often, the main manuscript only allows for an overview description of the methods. Use the supplement to describe all your methods, models and approaches in a lot of detail. Reference specific parts of your code as needed.

{{< pagebreak >}}

# Additional results

Show additional results here. Those can be some useful exploratory/descriptive figures or tables, or results from additional analyses that didn't make it into the main text.

## Example additional result

@tbl-resulttable1 shows an additional table summarizing a model fit.

```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Another fit table."
#| echo: FALSE
#resulttable1 = readRDS(here("results","resulttable1.rds"))
#knitr::kable(resulttable1)
```

@fig-result2 shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result2
#| fig-cap: "Height and weight."
#| echo: FALSE
#knitr::include_graphics(here("results","height_weight.png"))
```

{{< pagebreak >}}

# Discussion

Any additional discussion regarding the supplementary material/findings.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

{{< pagebreak >}}

# References

### 2023-03-13 products/slides/Slides.qmd
removed unnecessary files

---
title: Example Quarto slides
author: Andreas Handel
format: revealjs
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

## Overview

-   A few simple slides using the [`revealjs` (html) format](https://quarto.org/docs/presentations/revealjs/).
-   For other formats (e.g. Powerpoint, or Beamer/pdf), see [here](https://quarto.org/docs/guide/).

## Example slide

This shows the summary table. It is pulled in as R chunk.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
#resulttable=readRDS(here::here("results","summarytable.rds"))
#knitr::kable(resulttable)
```

## Example slide

This shows a figure created by the analysis script. It is inserted using Markdown (not knitr code, but that would be possible too).

![](images/height_weight.png){fig-align="center" width="420"}

## Example slide

This shows the model fitting results as table.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
#resulttable2 = readRDS(here::here("results","resulttable2.rds"))
#knitr::kable(resulttable2)
```

## Example slide with reference

This paper [@leek2015] discusses types of analyses.

## References

### 2023-03-14 code/analysis_code/exploratory/exploratory_analysis.qmd
general cleaning

--- 2023-03-13 docs/code/analysis_code/exploratory_analysis_LML.qmd
@@
 author: "Leah Lariscy"
-format: html
 editor: visual
+format:
+  html:
+    toc: true
+    toc-depth: 3
 theme: solar
@@
 n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))
+prop_pos_bio <- readRDS(here("data/processed_data/prop_pos_bio.rds"))
+prop_pos_sample <- readRDS(here("data/processed_data/prop_pos_sample.rds"))
 
@@
 
-### Case 7 day moving average overtime, bar plot and GAM model
+### Case 7 day moving average overtime
 
@@
 
-### PCR positives per day over time, bar plot and GAM model
+### PCR positives per day over time
 
@@
   geom_bar(stat = "identity", color = '#FF7E5A') +
-  stat_smooth(method = "gam", color = 'black') +
+  geom_line() +
   ylab(label = "Clinical Case 14-Day Moving Average") +
@@
 
-### Log10 total copies in wastewater over time (avg N1/N2)
+### Log10 viral load in wastewater over time (avg N1/N2)
 
@@
 
-### Log10 viral load in wastewater for N1 and N2
+### Log10 viral load in wastewater (both N1 and N2)
 
@@
 
+### Percent positive technical replicates per sample over time
+
+```{r}
+prop_pos_sample %>% group_by(date) %>% 
+  summarize(avg_pos_rate = mean(pos_rate_tr)) %>%
+
+  ggplot(aes(date, avg_pos_rate)) +
+  geom_point() +
+  geom_line()
+
+```
+
 ## Co-variate analysis: correlations between WW and clinical variables
 
-### Log10 copies per liter vs case 7DMA (N1/N2 average)
+### Log10 copies per liter vs log10 case 7DMA (N1/N2 average)
 
@@
 
-### Log10 copies per liter vs case 7DMA (N1 & N2)
+### Log10 copies per liter vs log10 case 7DMA (N1 & N2)
 
@@
 
-### Log10 Copies per Liter vs daily clinical cases
+### Log10 Copies per Liter vs log10 daily clinical cases
 
@@
 
-### Log10 Copies per Liter vs daily clinical cases (N1 N2 average)
+### Log10 Copies per Liter vs log10 daily clinical cases (N1 N2 average)
 
@@
 
-### Log10 viral load vs 7DMA
+### Log10 viral load vs log10 7DMA
 
@@
 
-### Log10 viral load vs 7DMA (N1 & N2)
+### Log10 viral load vs log10 7DMA (N1 & N2)
 
@@
 
-### Look at relationship of copies/L and 7dma over time
-
-```{r}
-all_data_combine %>% #filter(target == "N2") %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L), 
-            mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L),
-            mean_7dma = mean(mean_7dma)) %>% 
-  ggplot(aes(x=date)) +
-  #geom_point(aes(y=log10(sum_mean_copies_L)/mean_7dma)) + 
-  geom_line(aes(y=(log10(sum_mean_copies_L)/mean_7dma), color = "red")) +
-  geom_line(aes(y=log10(mean_7dma)), color = "blue")
-
-```
-
-### Log10 viral load vs daily positive cases
+### Log10 viral load vs log10 daily positive cases
 
@@
 
-### Log10 viral load vs daily positive cases
+### Log10 viral load vs log10 daily positive cases
 
@@
 ```
+
+### Sample percent positive vs log10 7DMA
+
+```{r}
+
+```

### 2023-03-14 code/processing_code/processing.qmd
general cleaning

--- 2023-02-24 code/processing_code/data_processing_LML.qmd
@@
 
+`wbe_year2` includes: date, facility, target, bio rep, collection number, ct, copies/uL/rxn, copies/L
+
 ```{r}
@@
 saveRDS(wbe_year2, here("data/processed_data/year2_cfx_clean.rds"))
+wbe_year2
 ```
@@
 ### Summarizing CFX data for downstream processing
+
+`wbe_year2_avg` - copies/L is averaged across biological replicates
 
@@
 saveRDS(wbe_year2_avg, here("data/processed_data/year2_cfx_avg.rds"))
+wbe_year2_avg
 ```
@@
 saveRDS(plant_clean, here("data/processed_data/year2_plant_clean.rds"))
+plant_clean
 ```
@@
 saveRDS(n1_n2_plant, here("data/processed_data/year2_cfx_plant_combined.rds"))
+n1_n2_plant
 ```
@@
 saveRDS(covid, here("data/processed_data/all_covid_combined.rds"))
+covid
 ```
@@
 saveRDS(all_data, here("data/processed_data/year2_all_data_combined.rds"))
+all_data
 ```
@@
 ```
+
+### Calculate percent positive of technical reps
+
+This isn't super important, but I'm including anyways.
+
+This will calculate the percent of positive technical replicates per biological replicate
+
+```{r}
+wbe_tr <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
+  dplyr::summarise(
+    n=n(), 
+    n.miss=sum(is.na(ct)), 
+    ct.mean=mean(ct,na.rm=T), 
+    ct.sd=sd(ct,na.rm=T)
+  ) %>% 
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(n.pos = n - n.miss) %>% 
+  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
+  mutate(percent_pos = (n.pos/n)*100)
+wbe_tr
+```
+
+### Calculate percent positive per biological replicate
+
+`pos_rate_tr` will calculate the percento of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)
+
+`percent_pos_bio` will calculate percent of positive biological replicates for each target
+
+```{r}
+wbe_br <- wbe_tr %>%
+  group_by(date, facility, target) %>% 
+  dplyr::summarise(
+    n.bio = n(),
+    n.bio.non.miss = sum(!is.na(ct.mean)),
+    n.bio.miss = sum(is.na(ct.mean)),
+    n.total = sum(n), 
+    n.total.miss = sum(n.miss), 
+    bio.ct.mean = mean(ct.mean, na.rm = T), 
+    bio.ct.sd = sd(ct.mean, na.rm=T), 
+    pos_rate_tr = mean(percent_pos),
+    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
+  ) %>%
+  mutate(percent_pos_bio = (n.bio.non.miss/n.bio)*100) %>% 
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup()
+wbe_br
+saveRDS(wbe_br, here("data/processed_data/prop_pos_bio.rds"))
+```
+
+### Calculate percent positive per sample
+
+`pos_rate_tr` calculates the percent of positive technical replicates, averaged across all biological replicates and averaged across both targets
+
+`prop.pos` calculates percent of positive targets for each sample
+
+```{r}
+wbe_samp <- wbe_br %>%
+  group_by(date, facility) %>% 
+  dplyr::summarise(
+    n.pos = sum(!is.na(bio.ct.mean)),
+    n.miss = sum(is.na(bio.ct.mean)),
+    pos_rate_tr = mean(pos_rate_tr)) %>%
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(prop.pos = n.pos/(n.pos+n.miss)*100)
+wbe_samp
+saveRDS(wbe_samp, here("data/processed_data/prop_pos_sample.rds"))
+```

### 2023-03-17 code/modeling_code/VL_7DMA_model.Rmd
model building

---
title: "VL_7DMA_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Waste water viral load

Outcome: Clinical case seven day moving average

## Load packages and data

```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set VL as predictor and 7DMA as outcome

```{r}
recipe_7dma <- recipe(pcr_pos_7dma ~ total_copies, data = train_data)
```

## Create workflow

```{r}
workflow_7dma <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_7dma)
```

## Create model using workflow

```{r}
fit_7dma <- workflow_7dma %>% 
  fit(data = train_data)

tidy(fit_7dma)
```

## Use model on the test data

```{r}
aug_test <- augment(fit_7dma, test_data)
aug_test %>% select(pcr_pos_7dma, .pred)
```

## Use RMSE to evaluate model performance

```{r}
aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
  geom_point()
```

### 2023-03-17 code/modeling_code/data_splitting.Rmd
model building

---
title: "Data splitting"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

```{r}
library(tidyverse) 
library(tidymodels)
library(here)
library(skimr)
```

## Data splitting for model building

I am first going to read in my final data frame that I already processed in the processing script.

```{r}
covid_ww_data <- readRDS(here("data/processed_data/final_combined_data.RDS"))
covid_ww_data <- covid_ww_data %>% drop_na()
tibble(covid_ww_data)
```

Since our data set is large enough, we can split it into a training set and a testing set.

First though, I want to split the data such that one set contains observations prior to July 2022 and the latter contains observations after July 2022. From my exploratory analysis, namely, comparing temporal trends of wastewater viral load with clinical case moving averages, I noted that there appears to be a disconnect in trends after the Omicron BA.5 wave (Late summer/fall 2022). As such, I would like to build models using data prior to this time and then test the model on the latter portion of the data to see if there is a difference.

### Splitting by date

```{r}
train_dates <- covid_ww_data %>% filter(collection_num <= 193)

test_dates <- covid_ww_data %>% filter(collection_num > 193)

saveRDS(test_dates, here(data/modeling_data/test_dates.RDS))
```

### Splitting by proportion for model training and testing

```{r}
data_split <- initial_split(train_dates, prop = 3/4)

train_data <- training(data_split)
test_data <- testing(data_split)
```

### 2023-03-17 code/processing_code/processing.qmd
model building

@@
     pos_rate_tr = mean(pos_rate_tr)) %>%
-  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
-  ungroup() %>% 
-  mutate(prop.pos = n.pos/(n.pos+n.miss)*100)
+   mutate(prop.pos = n.pos/(n.pos+n.miss)*100) %>% 
+  #mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup()
+
 wbe_samp
@@
 ```
+
+### Combine all data: wastewater, clinical cases, percent positive
+
+```{r}
+final_data <- full_join(all_data, wbe_samp, by = c("date", "facility"))
+saveRDS(final_data, here("data/processed_data/final_combined_data.RDS"))
+```

### 2023-03-17 code/modeling_code/VL_7DMA_model.Rmd
update manuscript

@@
 ```{r}
+set.seed(626)
 fit_7dma <- workflow_7dma %>% 
@@
 
-## Use RMSE to evaluate model performance
+## Use RMSE and RSQ to evaluate model performance
 
 ```{r}
-aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
+VL_7DMA_metrics <- full_join(rmse, rsq)
+
+saveRDS(VL_7DMA_metrics, here("data/modeling_data/metrics/VL_7DMA_performance.RDS"))
+
 ```
@@
 aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
+  geom_point() +
+  stat_smooth(method = "lm")
+ggsave(here("data/modeling_data/metrics/VL_7DMA_predict.png"))
+```
+
+## Plot residual values
+
+```{r}
+aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
+aug_residual %>% ggplot(aes(.pred, residual)) +
   geom_point()
+
+ggsave(here("data/modeling_data/metrics/VL_7DMA_residual.png"))
 ```

### 2023-03-17 code/modeling_code/data_splitting.Rmd
update manuscript

@@
 ```{r}
+set.seed(420)
 train_dates <- covid_ww_data %>% filter(collection_num <= 193)
@@
 
-saveRDS(test_dates, here(data/modeling_data/test_dates.RDS))
+saveRDS(test_dates, here("data/modeling_data/test_dates.RDS"))
 ```
@@
 test_data <- testing(data_split)
+
+saveRDS(train_data, here("data/modeling_data/training_data.RDS"))
+saveRDS(test_data, here("data/modeling_data/testing_data.RDS"))
 ```

### 2023-03-17 products/manuscript/Manuscript.qmd
update manuscript

@@
 # load a few R packages
-#library(here)
-#library(knitr)
+library(here)
+library(knitr)
 ```
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+_Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production [1,2]._
+
+_Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers [3]._
+
+_Wastewater-based detection methods have been utilized across the globe as an independent indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends [4].
+WBE methods may also become a primary tool in understanding transmission dynamics as case under reporting increases._
 
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+_24-hour composite samples collected twice weekly, stored at 4C until ready for extraction. 6 extraction replicates were performed for each collection date at treatment facility._
+
+_2019-nCoV N1 and N2 primer-probe sets were used for universal detection of SARS-CoV-2._
+
+_The concentration per reaction of each gene target was determined by the corresponding standard curve, then sample concentration in copies per liter was determined by Equation 1.
+Total daily copies per day per WRF was determined by Equation 2._
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
+_Is there a relationship between viral load and clinical case observations?_
+
+_Is there a relationship between qPCR percent positive and clinical case observations?_
+
+_What do viral load trends look like over time?_
+
+_What do clinical case trends look like over time?_
+
+_How do each of these temporal trends compare to one another over time?_
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+_data description_
+
+_cleaning process_
+
+_analysis process_
 
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+_Wastewater SARS-CoV-2 data was obtained through weekly surveillance efforts utilizing RT-qPCR methods._
+
+_Water Reclamation Facility data was obtained through data share directly from the WRF representatives._
+
+_COVID data downloaded from Georgia DPH website._
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+_qPCR data_
+
+_DPH COVID data_
 

### 2023-03-18 products/manuscript/Manuscript.qmd
update mansucript

--- 2023-03-13 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
+title: "Data Analysis Project"
 subtitle: ""
-author: Andreas Handel
+author: Leah Lariscy
 date: "`r Sys.Date()`"
@@
 
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
 # load a few R packages
-#library(here)
-#library(knitr)
+library(here)
+library(knitr)
 ```
@@
 # Summary/Abstract
-_Write a summary of your project._
+
+*Write a summary of your project.*
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.
+
+Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.
+
+Wastewater-based detection methods have been utilized across the globe as an independent indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends \[4\].
+WBE methods may also become a primary tool in understanding transmission dynamics as case under reporting increases.
 
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
+24-hour composite samples collected twice weekly, stored at 4C until ready for extraction. 6 extraction replicates were performed for each collection date at treatment facility.
+
+2019-nCoV N1 and N2 primer-probe sets were used for universal detection of SARS-CoV-2.
+
+The concentration per reaction of each gene target was determined by the corresponding standard curve, then sample concentration in copies per liter was determined by Equation 1.
+Total daily copies per day per WRF was determined by Equation 2.
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
+
+Is there a relationship between viral load and clinical case observations?
+
+Is there a relationship between qPCR percent positive and clinical case observations?
+
+What do viral load trends look like over time?
+
+What do clinical case trends look like over time?
+
+How do each of these temporal trends compare to one another over time?
 
@@
 Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+## Data aquisition
 
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Wastewater SARS-CoV-2 data was obtained through weekly surveillance efforts utilizing RT-qPCR methods.
+
+Water Reclamation Facility data was obtained through data share directly from the WRF representatives.
+
+COVID data downloaded from Georgia DPH website.
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+
+*qPCR data*
+
+WWTP data
+
+*DPH COVID data*
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
+Simple analyses: Wastewater and clinical data over time, linear regression model of wastewater vs clinical case data
+
+Full analyses: Wastewater data as predictor for clinical case outcomes
+
+Steps I took to build and tune model
 
@@
 ## Exploratory/Descriptive analysis
-
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
 
@@
 
+![Distribution of recorded seven-day moving average for each day](images/7DMA_histo.png)
+
+![Distribution of viral load per sample collection per target](images/total_copies_histo.png)
+
 ## Basic statistical analysis
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
+*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
 simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
 
@@
 
+![Clinical case seven-day moving average over time](images/7DMA_time_plot.png)
+
+![Wastewater viral load over time](images/avg_total_copies_time.png)
+
+![Linear regression of viral load vs case seven-day moving average](images/total_copies_avg_vs_7DMA.png)
+
 ## Full analysis
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
+*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
 This might again be code that is best placed in one or several separate R scripts that need to be well documented.
 You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+Then you load them here.*
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+
+*Summarize what you did, what you found and what it means.*
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+*Discuss what you perceive as strengths and limitations of your analysis.*
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*What are the main take-home messages?*
+
+*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
 

### 2023-03-18 code/processing_code/processing.qmd
update manuscript and readme

@@
   ) %>%
-  select(date, cases.symptom.onset, cases.reported, pcr_tests, pcr_pos, pcr_pos_7dma, pcr_pos_14dma) 
-  #full_join(cases.wrf, by = "date")
+  select(date, cases.symptom.onset, cases.reported, pcr_tests, pcr_pos, pcr_pos_7dma, pcr_pos_14dma) %>%
+
+  mutate(prop_pos = pcr_pos/pcr_tests)
 

### 2023-03-18 products/manuscript/Manuscript.qmd
update manuscript and readme

@@
 format:
-  docx:
+  html:
     toc: false
@@
 library(knitr)
+library(tidyverse)
+library(skimr)
 ```
@@
 Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.
-
 Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.
-
-Wastewater-based detection methods have been utilized across the globe as an independent indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends \[4\].
-WBE methods may also become a primary tool in understanding transmission dynamics as case under reporting increases.
+Wastewater-based detection methods have been utilized across the globe as an independent and parallel indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends.
+WBE methods may also become a primary tool in understanding transmission dynamics as case under-reporting increases.
 
@@
 
-24-hour composite samples collected twice weekly, stored at 4C until ready for extraction. 6 extraction replicates were performed for each collection date at treatment facility.
+### Wastewater SARS-CoV-2 surveillance data
 
-2019-nCoV N1 and N2 primer-probe sets were used for universal detection of SARS-CoV-2.
+24-hour composite wastewater samples were collected twice weekly from three treatment facilities in Clarke County and refrigerated until ready for extraction.
+RNA extractions were performed for each sample in replicates (n=6).
+Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) and SARS-CoV-2 specific primer-probe pairs were used to amplify each RNA extraction replicate in technical replicates (n=3).
+The concentration per reaction of each gene target was determined by the amplification standard curve from positive controls corresponding to the targets.
 
-The concentration per reaction of each gene target was determined by the corresponding standard curve, then sample concentration in copies per liter was determined by Equation 1.
-Total daily copies per day per WRF was determined by Equation 2.
+Wastewater influent flow data, including total daily flow volume and total suspended solids rate, were collected for each corresponding sample date.
+This was used to estimate the viral load of SARS-CoV-2, also known as total daily copies.
+
+### COVID-19 testing data
+
+COVID-19 data sets were downloaded from the Georgia Department of Public Health website (accessed Jan 2023), of which a subset was selected which includes symptom onset, PCR test, and PCR positive test data for each date since the first case was detected.
 
@@
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
 {{< pagebreak >}}
@@
 
-## Data aquisition
+## Data acquisition
 
-Wastewater SARS-CoV-2 data was obtained through weekly surveillance efforts utilizing RT-qPCR methods.
+**Wastewater SARS-CoV-2 viral load data** was obtained through weekly surveillance efforts utilizing RT-qPCR methods (Section 2.2).
 
-Water Reclamation Facility data was obtained through data share directly from the WRF representatives.
+**WWTP flow data** was obtained through data share directly from plant representatives.
 
-COVID data downloaded from Georgia DPH website.
+**COVID-19 case data** was downloaded from Georgia Department of Public Health's COVID-19 Status Report Page in January 2023 \[x\].
 
@@
 
-*qPCR data*
+### RT-qPCR data
 
-WWTP data
+For each target, viral copies per microliter (cp/uL) of each PCR reaction was estimated by transforming the cycle threshold value (Ct) using the appropriate standard curve equation.
+For reactions with no Ct value (no detection), cp/uL was set to the theoretical limit of detection (LOD).
+From this, N1 and N2 data frames were combined and copies per liter (cp/L) of sample for both targets was estimated using Equation 1.
 
-*DPH COVID data*
+```{r, echo=FALSE, message=FALSE}
+
+readRDS(here("data/processed_data/year2_cfx_clean.rds")) %>% glimpse()
+
+```
+
+### Treatment plant flow data
+
+For each collection date, corresponding , were combined with the qPCR data.
+Viral load (total daily viral copies) was then estimated by multiplying concentration estimates by daily flow volume (Equation 2).
+
+```{r, echo=FALSE, message=FALSE}
+readRDS(here("data/processed_data/year2_cfx_plant_combined.rds")) %>% glimpse()
+```
+
+### COVID testing data
+
+Data frames were combined into one and filtered to include only data pertaining to Clarke County. Proportion of positive tests per day was calculated by dividing PCR positive tests by total PCR tests.
+
+```{r, echo=FALSE, message=FALSE}
+readRDS(here("data/processed_data/all_covid_combined.rds")) %>% glimpse()
+```
 
@@
 
-Simple analyses: Wastewater and clinical data over time, linear regression model of wastewater vs clinical case data
+### Exploratory analysis
 
-Full analyses: Wastewater data as predictor for clinical case outcomes
+Wastewater and clinical data over time, linear regression model of wastewater vs clinical case data
 
-Steps I took to build and tune model
+### Predicting clinical case levels with wastewater surveillance data
+
+Wastewater data as predictor for clinical case outcomes
 
@@
 
-## Exploratory/Descriptive analysis
+## Descriptive analysis
 
@@
 
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
 ```{r}
@@
 
+## Basic statistical analysis
+
 ![Linear regression of viral load vs case seven-day moving average](images/total_copies_avg_vs_7DMA.png)
@@
 ## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
 

### 2023-03-20 code/modeling_code/PP_14DMA_model.Rmd
model building

---
title: "PP_14DMA_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Percent positive technical replicates per sample

Outcome: Clinical case 14 day moving average

## Load packages and data

```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set PP as predictor and 14dma as outcome

```{r}
recipe_14dma <- recipe(pcr_pos_14dma ~ pos_rate_tr, data = train_data)
```

## Create workflow

```{r}
workflow_14dma <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_14dma)
```

## Create model using workflow

```{r}
set.seed(626)
fit_14dma <- workflow_14dma %>% 
  fit(data = train_data)

tidy(fit_14dma)
```

## Use model on the training data

```{r}
aug_test <- augment(fit_14dma, train_data)
aug_test %>% select(pcr_pos_14dma, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = pcr_pos_14dma, .pred)
rsq <- aug_test %>% rsq(truth = pcr_pos_14dma, .pred)
PP_14dma_metrics <- full_join(rmse, rsq)

saveRDS(PP_14dma_metrics, here("data/modeling_data/metrics/PP_14dma_performance.RDS"))

PP_14dma_metrics

```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(pcr_pos_14dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/PP_14dma_predict.png"))
```

## Plot residual values

```{r}
aug_residual <- aug_test %>% mutate(residual = pcr_pos_14dma - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

ggsave(here("data/modeling_data/metrics/PP_14dma_residual.png"))
```

### 2023-03-20 code/modeling_code/PP_7DMA_model.Rmd
model building

--- 2023-03-17 code/modeling_code/VL_7DMA_model.Rmd
@@
 ---
-title: "VL_7DMA_model"
+title: "PP_7DMA_model"
 author: "Leah Lariscy"
@@
 
-Predictor: Waste water viral load
+Predictor: Percent positive technical replicates per sample
 
@@
 
-This is where we will set VL as predictor and 7DMA as outcome
+This is where we will set PP as predictor and 7DMA as outcome
 
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ total_copies, data = train_data)
+recipe_7dma <- recipe(case_7dma ~ pos_rate_tr, data = train_data)
 ```
@@
 
-## Use model on the test data
+## Use model on the training data
 
 ```{r}
-aug_test <- augment(fit_7dma, test_data)
-aug_test %>% select(pcr_pos_7dma, .pred)
+aug_test <- augment(fit_7dma, train_data)
+aug_test %>% select(case_7dma, .pred)
 ```
 
-## Use RMSE to evaluate model performance
+## Use RMSE and RSQ to evaluate model performance
 
 ```{r}
-aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
+PP_7DMA_metrics <- full_join(rmse, rsq)
+
+saveRDS(PP_7DMA_metrics, here("data/modeling_data/metrics/PP_7DMA_performance.RDS"))
+
+PP_7DMA_metrics
+
 ```
@@
 ```{r}
-aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
+aug_test %>% ggplot(aes(case_7dma, .pred)) +
+  geom_point() +
+  stat_smooth(method = "lm")
+ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
+```
+
+## Plot residual values
+
+```{r}
+aug_residual <- aug_test %>% mutate(residual = case_7dma - .pred)
+aug_residual %>% ggplot(aes(.pred, residual)) +
   geom_point()
+
+ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
 ```

### 2023-03-20 code/modeling_code/PP_DailyCase_model.Rmd
model building

---
title: "PP_DailyCase_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Percent positive technical replicates per sample

Outcome: Clinical case seven day moving average

## Load packages and data

```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set PP as predictor and DailyCase as outcome

```{r}
recipe_DailyCase <- recipe(pcr_pos ~ pos_rate_tr, data = train_data)
```

## Create workflow

```{r}
workflow_DailyCase <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_DailyCase)
```

## Create model using workflow

```{r}
set.seed(626)
fit_DailyCase <- workflow_DailyCase %>% 
  fit(data = train_data)

tidy(fit_DailyCase)
```

## Use model on the training data

```{r}
aug_test <- augment(fit_DailyCase, train_data)
aug_test %>% select(pcr_pos, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = pcr_pos, .pred)
rsq <- aug_test %>% rsq(truth = pcr_pos, .pred)
PP_DailyCase_metrics <- full_join(rmse, rsq)

saveRDS(PP_DailyCase_metrics, here("data/modeling_data/metrics/PP_DailyCase_performance.RDS"))

PP_DailyCase_metrics

```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(pcr_pos, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/PP_DailyCase_predict.png"))
```

## Plot residual values

```{r}
aug_residual <- aug_test %>% mutate(residual = pcr_pos - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

ggsave(here("data/modeling_data/metrics/PP_DailyCase_residual.png"))
```

### 2023-03-20 code/modeling_code/PP_PosTest_model.Rmd
model building

---
title: "PP_PosTest_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Percent positive technical replicates per sample

Outcome: Clinical case seven day moving average

## Load packages and data

```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set PP as predictor and PosTest as outcome

```{r}
recipe_PosTest <- recipe(percent_pos_test ~ pos_rate_tr, data = train_data)
```

## Create workflow

```{r}
workflow_PosTest <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_PosTest)
```

## Create model using workflow

```{r}
set.seed(626)
fit_PosTest <- workflow_PosTest %>% 
  fit(data = train_data)

tidy(fit_PosTest)
```

## Use model on the training data

```{r}
aug_test <- augment(fit_PosTest, train_data)
aug_test %>% select(percent_pos_test, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
PP_PosTest_metrics <- full_join(rmse, rsq)

saveRDS(PP_PosTest_metrics, here("data/modeling_data/metrics/PP_PosTest_performance.RDS"))

PP_PosTest_metrics

```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(percent_pos_test, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/PP_PosTest_predict.png"))
```

## Plot residual values

```{r}
aug_residual <- aug_test %>% mutate(residual = percent_pos_test - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

ggsave(here("data/modeling_data/metrics/PP_PosTest_residual.png"))
```

### 2023-03-20 code/modeling_code/data_splitting.Rmd
model building

@@
 ```{r}
-covid_ww_data <- readRDS(here("data/processed_data/final_combined_data.RDS"))
-covid_ww_data <- covid_ww_data %>% drop_na()
+covid_ww_data <- readRDS(here("data/processed_data/final_sum_combined_data.RDS"))
+#covid_ww_data <- covid_ww_data %>% drop_na()
 tibble(covid_ww_data)

### 2023-03-20 code/modeling_code/log10VL_7DMA_model.Rmd
model building

--- 2023-03-17 code/modeling_code/VL_7DMA_model.Rmd
@@
 ---
-title: "VL_7DMA_model"
+title: "log10VL_7DMA_model"
 author: "Leah Lariscy"
@@
 
-Predictor: Waste water viral load
+Predictor: log10 wastewater viral load
 
@@
 
-This is where we will set VL as predictor and 7DMA as outcome
+This is where we will set log10VL as predictor and 7DMA as outcome
 
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ total_copies, data = train_data)
+recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL, data = train_data)
 ```
@@
 
-## Use model on the test data
+## Use model on the training data
 
 ```{r}
-aug_test <- augment(fit_7dma, test_data)
+aug_test <- augment(fit_7dma, train_data)
 aug_test %>% select(pcr_pos_7dma, .pred)
@@
 rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
-VL_7DMA_metrics <- full_join(rmse, rsq)
+log10VL_7DMA_metrics <- full_join(rmse, rsq)
 
-saveRDS(VL_7DMA_metrics, here("data/modeling_data/metrics/VL_7DMA_performance.RDS"))
+saveRDS(log10VL_7DMA_metrics, here("data/modeling_data/metrics/log10VL_7DMA_performance.RDS"))
+
+log10VL_7DMA_metrics
 
@@
   stat_smooth(method = "lm")
-ggsave(here("data/modeling_data/metrics/VL_7DMA_predict.png"))
+ggsave(here("data/modeling_data/metrics/log10VL_7DMA_predict.png"))
 ```
@@
 
-ggsave(here("data/modeling_data/metrics/VL_7DMA_residual.png"))
+ggsave(here("data/modeling_data/metrics/log10VL_7DMA_residual.png"))
 ```

### 2023-03-20 code/modeling_code/log10VL_PP_7DMA_model.Rmd
model building

--- 2023-03-20 code/modeling_code/log10VL_7DMA_model.Rmd
@@
 ---
-title: "log10VL_7DMA_model"
+title: "log10VL_PP_7DMA_model"
 author: "Leah Lariscy"
@@
 
-Predictor: log10 wastewater viral load
+Predictor: Waste water viral load
 
@@
 
-This is where we will set log10VL as predictor and 7DMA as outcome
+This is where we will set VL as predictor and 7DMA as outcome
 
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL, data = train_data)
+recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL + pos_rate_tr, data = train_data)
 ```
@@
 rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
-log10VL_7DMA_metrics <- full_join(rmse, rsq)
+log10VL_PP_7DMA_metrics <- full_join(rmse, rsq)
 
-saveRDS(log10VL_7DMA_metrics, here("data/modeling_data/metrics/log10VL_7DMA_performance.RDS"))
+saveRDS(log10VL_PP_7DMA_metrics, here("data/modeling_data/metrics/log10VL_PP_7DMA_performance.RDS"))
 
-log10VL_7DMA_metrics
+log10VL_PP_7DMA_metrics
 
@@
   stat_smooth(method = "lm")
-ggsave(here("data/modeling_data/metrics/log10VL_7DMA_predict.png"))
+ggsave(here("data/modeling_data/metrics/log10VL_PP_7DMA_predict.png"))
 ```
@@
 
-ggsave(here("data/modeling_data/metrics/log10VL_7DMA_residual.png"))
+ggsave(here("data/modeling_data/metrics/log10VL_PP_7DMA_residual.png"))
 ```

### 2023-03-20 code/processing_code/processing.qmd
model building

@@
 ```{r}
-final_data <- full_join(all_data, wbe_samp, by = c("date", "facility"))
+final_data <- full_join(all_data, wbe_samp, by = c("date", "facility")) %>% 
+  mutate(log10_VL = log10(total_copies), percent_pos_test = prop_pos*100) %>% 
+  drop_na(facility)
+
 saveRDS(final_data, here("data/processed_data/final_combined_data.RDS"))
+
+final_data
 ```
+
+### Summarize data across treatment facilities then average across targets
+
+Group by: Date, Collection number
+
+Keep: total copies, pos_rate_tr, log10 total copies, % positive tests
+
+```{r}
+final_data_sum <- final_data %>% group_by(date, collection_num, target) %>% 
+  summarize(sum_total_copies = sum(total_copies),
+            pos_rate_tr = mean(pos_rate_tr),
+            percent_pos_test = mean(percent_pos_test),
+            case_7dma = mean(pcr_pos_7dma)) %>% 
+  ungroup(target) %>% 
+  summarize(avg_VL = mean(sum_total_copies),
+            pos_rate_tr = mean(pos_rate_tr),
+            percent_pos_test = mean(percent_pos_test),
+            case_7dma = mean(case_7dma)) %>% 
+  ungroup() %>% 
+  mutate(log10_VL = log10(avg_VL))
+
+saveRDS(final_data_sum, here("data/processed_data/final_sum_combined_data.RDS"))
+```

### 2023-03-21 code/modeling_code/PP_7DMA_model.Rmd
model building

@@
 rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
-PP_7DMA_metrics <- full_join(rmse, rsq)
+huber <- aug_test %>% huber_loss(truth = case_7dma, .pred)
+PP_7DMA_metrics <- bind_rows(rmse, rsq, huber)
 
@@
 ```
+
+## Plot 7DMA and prediction temporally
+
+```{r}
+aug_test %>% ggplot(aes(x=date)) +
+  geom_point(aes(y=case_7dma), color = "blue") +
+  geom_line(aes(y=case_7dma), color = "blue") +
+  geom_point(aes(y=.pred), color = "green") +
+  geom_line(aes(y=.pred), color = "green")
+
+```
+
+## Use model on testing data
+
+```{r}
+aug_test2 <- augment(fit_7dma, test_data)
+aug_test2 %>% select(case_7dma, .pred)
+```
+
+```{r}
+aug_test2 %>% ggplot(aes(x=date)) +
+  geom_point(aes(y=case_7dma), color = "blue") +
+  geom_line(aes(y=case_7dma), color = "blue") +
+ geom_point(aes(y=.pred), color = "green") +
+  geom_line(aes(y=.pred), color = "green")
+```
+
+## Evaluate model performance on testing set
+
+```{r}
+rmse <- aug_test2 %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test2 %>% rsq(truth = case_7dma, .pred)
+huber <- aug_test2 %>% huber_loss(truth = case_7dma, .pred)
+PP_7DMA_test_metrics <- bind_rows(rmse, rsq, huber)
+
+saveRDS(PP_7DMA_test_metrics, here("data/modeling_data/metrics/PP_7DMA_test_performance.RDS"))
+
+PP_7DMA_test_metrics
+```
+
+## Use model on testing dates
+
+```{r}
+aug_test3 <- augment(fit_7dma, test_dates)
+aug_test3 %>% select(case_7dma, .pred)
+```
+
+## Evaluate model performance on testing dates
+
+```{r}
+rmse <- aug_test3 %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test3 %>% rsq(truth = case_7dma, .pred)
+huber <- aug_test3 %>% huber_loss(truth = case_7dma, .pred)
+PP_7DMA_BA5_metrics <- bind_rows(rmse, rsq, huber)
+
+saveRDS(PP_7DMA_BA5_metrics, here("data/modeling_data/metrics/PP_7DMA_BA5_performance.RDS"))
+
+PP_7DMA_BA5_metrics
+```
+
+```{r}
+aug_test3 %>% ggplot(aes(x=date)) +
+  geom_point(aes(y=case_7dma), color = "blue") +
+  geom_line(aes(y=case_7dma), color = "blue") +
+ geom_point(aes(y=.pred), color = "green") +
+  geom_line(aes(y=.pred), color = "green")
+```

### 2023-03-21 code/processing_code/processing.qmd
model building

@@
 
+n1 \<- n1 %\>% mutate(ct = as.numeric(ct),
+
+copy_num_uL_rxn = as.numeric(10\^((ct-36.046)/-3.5293)), #use standard curve slope here
+
+copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004), #replace NAs with limit of detection (0.004 copies per uL of the reaction)
+
+run_date = as.Date(run_date))
+
 ```{r}
@@
               copy_num_uL_rxn = as.numeric(10^((ct-36.046)/-3.5293)), #use standard curve slope here
-              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004)) #replace NAs with limit of detection (0.004 copies per uL of the reaction)
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004), #replace NAs with limit of detection (0.004 copies per uL of the reaction)
+              run_date = as.Date(run_date))
 
@@
               copy_num_uL_rxn = as.numeric(10^((ct-37.731)/-3.2505)),
-              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004),
+              run_date = as.Date(run_date))
 
@@
 
-wbe_year2 <- bind_rows(n1, n2) %>% #combine dataframes with same number of rows
+wbe_year2 <- bind_rows(n1, n2) %>% #combine dataframes with same variables
   mutate(
@@
 plant_clean <- plant %>% 
-  mutate(date = as.Date(date, format = "%m/%d/%Y"), #change to date format
-         influent_flow_L = influent_flow_mg*1e6*231*(0.0254^3)*1000) %>%  #convert flow 
+  mutate(date = as.Date(date), #change to date format
+         influent_flow_L = influent_flow_mg*1e6*231*(0.0254^3)*1000) %>%  #convert flow from millions of gallons to liters
   mutate(facility = as.factor(wrf), facility = recode(facility, NO = "A", MI = "B", CC = "C"), facility = ordered(facility, levels = c("A", "B", "C"))) %>% #set facility to factor and then set factor levels
@@
 ```{r}
-n1_n2_plant <- left_join(wbe_year2_avg, plant_clean, by = c("date","facility")) %>% 
+n1_n2_plant <- left_join(wbe_year2_avg, plant_clean, by = c("date", "facility")) %>% 
   mutate(total_copies = avg_copy_L * influent_flow_L)

### 2023-04-24 code/analysis_code/exploratory/exploratory_analysis.qmd
model building
fixing issues

@@
 
-```{r}
+```{r, message=FALSE}
 library(tidyverse)
@@
 
-## Load data frames generated in data processing script
+## Load final data frame generated in data processing script
 
 ```{r}
-here()
-n1_n2_clean <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
-plant_clean <- readRDS(here("data/processed_data/year2_plant_clean.rds"))
-n1_n2_plant_combine <- readRDS(here("data/processed_data/year2_cfx_plant_combined.rds"))
-covid_combine <- readRDS(here("data/processed_data/all_covid_combined.rds"))
-all_data_combine <- readRDS(here("data/processed_data/year2_all_data_combined.rds"))
-recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))
-n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))
-prop_pos_bio <- readRDS(here("data/processed_data/prop_pos_bio.rds"))
-prop_pos_sample <- readRDS(here("data/processed_data/prop_pos_sample.rds"))
+data_clean <- readRDS(here("data/processed_data/final_sum_combined_data.RDS"))
 
@@
 
-### Look at `all_data_combine`
+### Look at `data_clean`
 
 ```{r}
-tibble(all_data_combine)
+tibble(data_clean)
 ```
@@
 ### Variation analysis
-
-Graphs to make:
 
@@
 
-Graphs to make:
+-   Scatter plot of wastewater and clinical variables over time
 
-## Single variate analysis
+-   Linear regressions between wastewater and clinical variables
+
+## Variation analysis
 
@@
 ```{r}
-covid_combine %>% ggplot(aes(pcr_pos_7dma)) +
+data_clean %>% ggplot(aes((case_7dma))) +
   geom_histogram(fill = "#C06C84") +
@@
 ggsave(here("figures/exploratory/variate/7DMA_histo.png"))
-summary(covid_combine$pcr_pos_7dma)
+summary(data_clean$case_7dma)
 ```
+
+Slight bimodal distribution in 7DMA
 
@@
 ```{r}
-covid_combine %>% ggplot(aes(pcr_pos)) +
+data_clean %>% ggplot(aes(pcr_pos)) +
   geom_histogram(fill = '#F8B195') +
@@
 ggsave(here("figures/exploratory/variate/pos_cases_histo.png"))
-summary(covid_combine$pcr_pos)
+summary(data_clean$pcr_pos)
 ```
 
-### Clinical case 14 day moving average histogram
+The distribution of daily cases is heavily skewed to the right. Log-transforming gives a more normal distribution
+
+### Clinical case percent positive tests
 
 ```{r}
-covid_combine %>% ggplot(aes(pcr_pos_14dma)) +
+data_clean %>% ggplot(aes(percent_pos_test)) +
   geom_histogram(fill = "#FF7E5A")
 
-summary(covid_combine$pcr_pos_14dma)
+summary(data_clean$percent_pos_test)
 
@@
 
-### Total copies in WW at county level histogram
+There is also a slight bimodal distribution with percent of positive tests, but nonetheless the distribution is right-skewed.
+
+### Wastewater viral load histogram
 
 ```{r}
-sum_daily_copies_county <- all_data_combine %>% 
-  drop_na() %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_total_copies = mean(total_copies)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_total_copies = sum(mean_total_copies))
-
-  sum_daily_copies_county %>% ggplot(aes(log10(sum_total_copies), fill = target)) +
+    data_clean %>% ggplot(aes(avg_VL)) +
   geom_histogram() +
-      scale_fill_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-    xlab(label = "Log10 Total Daily County-Level Copies") +
+    xlab(label = "Wastewater viral load") +
     ylab(label = "Count") +
@@
 
-ggsave(here("figures/exploratory/variate/total_copies_histo.png"))
-  log10_sum_daily_copies <- log10(sum_daily_copies_county$sum_total_copies)
-  summary(log10_sum_daily_copies)
+ggsave(here("figures/exploratory/variate/VL_histo.png"))
+
 ```
 
-### Copies per liter histogram
+Wastewater viral load is also very right-skewed. We may need to log-transform some variables before doing more statistical analysis.
+
+### Log10 viral load histogram
 
 ```{r}
-all_data_combine %>% #filter(target == "N2") %>% 
-  drop_na() %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>%
+data_clean %>% 
+  ggplot(aes(log10_VL)) +
+  geom_histogram()
 
-  ggplot(aes(log10(sum_mean_copies_L), fill = target)) +
+ggsave(here("figures/exploratory/variate/log10VL_histo.png"))
+
+```
+
+### Assay percent positivity histogram
+
+```{r}
+  data_clean %>% ggplot(aes(pos_rate_tr)) +
   geom_histogram() +
-      scale_fill_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-    xlab(label = "Log10 Copies Per Liter") +
+    xlab(label = "qPCR assay positivity") +
     ylab(label = "Count") +
@@
 
-ggsave(here("figures/exploratory/variate/copies_L_histo.png"))
+ggsave(here("figures/exploratory/variate/assay_pos_histo.png"))
 
 ```
+
+Again with the bimodal distribution, even more so this time.
 
@@
 
-```{r}
-all_data_combine %>% group_by(date) %>% 
-  summarize(mean_7dma = mean(pcr_pos_7dma),
-            mean_pcr_pos = mean(pcr_pos)) %>% 
-  ggplot(aes(x = date, y = mean_7dma)) +
-  geom_point(color = '#C06C84') +
-  geom_line(color = '#C06C84') +
+```{r, message=FALSE}
+data_clean %>% 
+  ggplot(aes(x = date, y = case_7dma)) +
+  geom_line(size =1.5, color = '#4c005c') +
   xlab(label = "Date") +
@@
 ```{r}
-all_data_combine %>% group_by(date) %>% 
-  summarize(mean_pcr_pos = mean(pcr_pos)) %>% 
-  ggplot(aes(x = date, y = mean_pcr_pos)) +
-  geom_bar(stat = "identity", color = '#F8B195') +
-  geom_line(color = 'black') +
+data_clean %>% 
+  ggplot(aes(x = date, y = pcr_pos)) +
+  geom_line(color = '#4c005c', size = 2) +
   ylab(label = "Positive Cases Per Day") +
@@
 
-### Case 14 day moving average over time
+### Clinical case percent positive tests over time
 
 ```{r}
-all_data_combine %>% group_by(date) %>% 
-  summarize(mean_14_dma = mean(pcr_pos_14dma)) %>%
-
-   ggplot(aes(x = date, y = mean_14_dma)) +
-  geom_bar(stat = "identity", color = '#FF7E5A') +
-  geom_line() +
-  ylab(label = "Clinical Case 14-Day Moving Average") +
+data_clean %>% 
+   ggplot(aes(x = date, y = percent_pos_test)) +
+  geom_line(size = 1.5) +
+  ylab(label = "Percent of positive clinical tests") +
   xlab(label = "Date") +
@@
 
-### Log10 viral load in wastewater over time (avg N1/N2)
+### Wastewater viral load over time
 
 ```{r}
-all_data_combine %>% #filter(target == "N2") %>% 
-  drop_na(target) %>% #remove NAs from analysis
-  group_by(date, facility) %>% #this will allow me to average the two targets
-  summarize(mean_total_copies = mean(total_copies)) %>% #average across N1 and N2
-  ungroup(facility) %>% #need to do this so I can add the avg total copies of each WRF
-  summarize(sum_total_copies = sum(mean_total_copies)) %>% #total daily county copies
-  ggplot(aes(date, log10(sum_total_copies))) +
-  geom_point(color = "#80256D") +
-  geom_line(color = "#80256D") +
+data_clean %>% 
+  ggplot(aes(date, avg_VL)) +
+  geom_line(color = "#2bce48", size = 1.5) +
    xlab(label = "Date") +
-  ylab(label = "Log10 Total Daily County-Level Copies") +
+  ylab(label = "Total Daily County-Level Copies") +
   theme_light()
@@
 
-### Log10 viral load in wastewater (both N1 and N2)
+### Log10 wastewater viral load over time
 
 ```{r}
-all_data_combine %>% 
-  group_by(date, target, facility) %>% 
-  drop_na(target) %>% 
-  summarize(mean_total_copies = mean(total_copies)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_total_copies = sum(mean_total_copies)) %>%
+data_clean %>% 
+  ggplot(aes(date, log10_VL)) +
+  geom_line(color = "#2bce48", size = 1.5) +
+   xlab(label = "Date") +
+  ylab(label = "Log10 Total Daily County-Level Copies") +
+  theme_light()
 
-  ggplot(aes(date, log10(sum_total_copies), color = target)) +
-  geom_point() +
-  geom_line() +
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-  xlab(label = "Date") +
-  ylab(label = "Log10 County-Level Viral Load") +
-  theme_light() +
-  facet_wrap(vars(target),nrow = 2)
-
- ggsave(here("figures/exploratory/covariate/total_copies_time.png"))
-```
-
-### Log10 copies per liter across time (both N1 and N2)
-
-```{r}
-all_data_combine %>%
-  drop_na(target) %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L)) %>% 
-  ggplot(aes(date, log10(sum_mean_copies_L),color = target)) +
-geom_line() +
-  geom_point() + 
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-  xlab(label = "Date") +
-  ylab(label = "Log10 County-Level Copies Per Liter") +
-  theme_light() +
-    facet_wrap(vars(target),nrow = 2)
-
- ggsave(here("figures/exploratory/covariate/copies_L_time.png"))
-
+ggsave(here("figures/exploratory/covariate/log10VL_time.png"))
 ```
@@
 
-```{r}
-prop_pos_sample %>% group_by(date) %>% 
-  summarize(avg_pos_rate = mean(pos_rate_tr)) %>%
+```{r, message=FALSE}
+data_clean %>% 
+  ggplot(aes(date, pos_rate_tr)) +
+geom_line(size =1.5, color = '#634499') +
+  xlab(label = "Date") +
+  ylab(label = "Percent Positive N1 & N2 assays") +
+  theme_light()
 
-  ggplot(aes(date, avg_pos_rate)) +
-  geom_point() +
-  geom_line()
+ggsave(here("figures/exploratory/covariate/percent_pos_time_plot.png"))
 
@@
 
+### Tests administered over time
+
+```{r}
+data_clean %>% ggplot(aes(date, test_admin)) +
+  geom_line(size = 1.5) +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/test_admin_time.png"))
+```
+
+Tests administered drop off significantly after the beginning of 2022. In all other time series, there is a spike in mid-2022, but that spike is missing here.
+
+From here on, I will split the data before and after Februrary 28, 2022 to see how correlations change.
+
 ## Co-variate analysis: correlations between WW and clinical variables
 
-### Log10 copies per liter vs log10 case 7DMA (N1/N2 average)
+### Log10 wastewater viral load vs case 7DMA
 
 ```{r}
-all_data_combine %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L), 
-            mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L),
-            mean_7dma = mean(mean_7dma)) %>% 
-  ggplot(aes(log10(sum_mean_copies_L), log10(mean_7dma))) +
+#full dataset
+data_clean %>% 
+  ggplot(aes(log10_VL, case_7dma)) +
   geom_point(color = "#80256D") +
@@
 
-ggsave(here("figures/exploratory/covariate/copies_L_avg_vs_7DMA.png"))
-```
+ggsave(here("figures/exploratory/covariate/log10VL_vs_7DMA.png"))
 
-### Log10 copies per liter vs log10 case 7DMA (N1 & N2)
-
-```{r}
-all_data_combine %>%  
-  drop_na(target) %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L), 
-            mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L),
-            mean_7dma = mean(mean_7dma)) %>% 
-  ggplot(aes(log10(sum_mean_copies_L), log10(mean_7dma), color = target)) +
-  geom_point() +
-  stat_poly_line() +
-  stat_poly_eq() +
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
+#data split before FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>% 
+  ggplot(aes(log10_VL, case_7dma)) +
+  geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
+  stat_poly_eq(color = "#80256D") +
    xlab(label = "Log10 County-Level Copies Per Liter") +
@@
 
-ggsave(here("figures/exploratory/covariate/copies_L_vs_7DMA.png"))
-```
+ggsave(here("figures/exploratory/covariate/log10VL_vs_7DMA_Y1.png"))
 
-### Log10 Copies per Liter vs log10 daily clinical cases
-
-```{r}
-all_data_combine %>%  
-  drop_na(target) %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L), 
-            mean_pcr_pos = mean(pcr_pos)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L),
-            mean_pcr_pos = mean_pcr_pos) %>% 
-  ggplot(aes(log10(sum_mean_copies_L), log10(mean_pcr_pos), color = target)) +
-  geom_point() +
-  stat_poly_line() +
-  stat_poly_eq() +
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-  xlab(label = "Log10 County-Level Copies Per Liter") +
-  ylab(label = "Log10 Daily Positive Clinical Cases") +
-  theme_light()
-
-ggsave(here("figures/exploratory/covariate/copies_L_vs_pcr_pos.png"))
-```
-
-### Log10 Copies per Liter vs log10 daily clinical cases (N1 N2 average)
-
-```{r}
-all_data_combine %>%  
-  drop_na(target) %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_copies_L = mean(avg_copy_L), 
-            mean_pcr_pos = mean(pcr_pos)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies_L = sum(mean_copies_L),
-            mean_pcr_pos = mean_pcr_pos) %>% 
-  ggplot(aes(log10(sum_mean_copies_L), log10(mean_pcr_pos))) +
+#data split after FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>% 
+  ggplot(aes(log10_VL, case_7dma)) +
   geom_point(color = "#80256D") +
@@
    xlab(label = "Log10 County-Level Copies Per Liter") +
-  ylab(label = "Log10 Daily Positive Clinical Cases") +
+  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
   theme_light()
 
-ggsave(here("figures/exploratory/covariate/copies_L_avg_vs_pcr_pos.png"))
+ggsave(here("figures/exploratory/covariate/log10VL_vs_7DMA_Y2.png"))
 ```
 
-### Log10 viral load vs log10 7DMA
+RSQ value does not change between full data and the subset of dates
+
+### Log10 wastewater viral load vs tests administered
 
 ```{r}
-all_data_combine %>% #filter(target == "N2") %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_daily_copies = mean(total_copies), 
-            mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies = sum(mean_daily_copies),
-            mean_7dma = mean(mean_7dma)) %>% 
-  ggplot(aes(log10(sum_mean_copies), log10(mean_7dma))) +
+#full dataset
+data_clean %>%  
+  ggplot(aes(log10_VL, test_admin)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 viral load") +
+  ylab(label = "PCR tests administered") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_testadmin.png"))
+
+#data split before FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>% 
+  ggplot(aes(log10_VL, test_admin)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 viral load") +
+  ylab(label = "PCR tests administered") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_testadmin_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>% 
+  ggplot(aes(log10_VL, test_admin)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 viral load") +
+  ylab(label = "PCR tests administered") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_testadmin_Y2.png"))
+```
+
+Correlations overall are weak, but the strongest positive correlation is found in the earlier dates
+
+### Log10 wastewater viral load vs percent of positive tests
+
+```{r}
+#full dataset
+data_clean %>%  
+  ggplot(aes(log10_VL, percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 wastewater viral load") +
+  ylab(label = "Percent of positive clinical cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_percentcase.png"))
+
+#data split before FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(log10_VL, percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 wastewater viral load") +
+  ylab(label = "Percent of positive clinical cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_percentcase_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(log10_VL, percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "Log10 wastewater viral load") +
+  ylab(label = "Percent of positive clinical cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_percentcase_Y2.png"))
+```
+
+RSQ is slightly lower in later dates than in full data set, but does not change in earlier dates
+
+### Log10 wastewater viral load vs daily reported cases
+
+```{r}
+#full dataset
+data_clean %>%  
+  ggplot(aes(log10_VL, log10(pcr_pos))) +
   geom_point(color = "#80256D") +
@@
   stat_poly_eq(color = "#80256D") +
-  xlab(label = "Log10 Daily Viral Load") +
-  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
+  xlab(label = "Log10 wastwater viral load") +
+  ylab(label = "Daily Positive Clinical Cases") +
   theme_light()
 
-ggsave(here("figures/exploratory/covariate/avg_total_copies_vs_7DMA.png"))
+ggsave(here("figures/exploratory/covariate/log10VL_dailycases.png"))
+
+#data split before FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(log10_VL, log10(pcr_pos))) +
+  geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
+  stat_poly_eq(color = "#80256D") +
+  xlab(label = "Log10 wastwater viral load") +
+  ylab(label = "Daily Positive Clinical Cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_dailycases_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>%  
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(log10_VL, log10(pcr_pos))) +
+  geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
+  stat_poly_eq(color = "#80256D") +
+  xlab(label = "Log10 wastwater viral load") +
+  ylab(label = "Daily Positive Clinical Cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/log10VL_dailycases_Y2.png"))
 ```
 
-### Log10 viral load vs log10 7DMA (N1 & N2)
+RSQ is higher is the data splits than in the full set. I'm not sure what to think of this, but it could be due to the high error associated with this metric.
+
+### Assay positivity vs 7DMA
 
 ```{r}
-all_data_combine %>% 
-  drop_na() %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_daily_copies = mean(total_copies), 
-            mean_7dma = mean(pcr_pos_7dma)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies = sum(mean_daily_copies),
-            mean_7dma = mean(mean_7dma)) %>% 
-  ggplot(aes(log10(sum_mean_copies), log10(mean_7dma), color = target)) +
+#full dataset
+data_clean %>% 
+  ggplot(aes(pos_rate_tr, case_7dma)) +
+  geom_point(color = "#00b0ba") +
+  stat_poly_line(color = "#00b0ba") +
+  stat_poly_eq(color = "black") +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_vs_7DMA.png"))
+
+#data split before FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(pos_rate_tr, case_7dma)) +
+  geom_point(color = "#00b0ba") +
+  stat_poly_line(color = "#00b0ba") +
+  stat_poly_eq(color = "black") +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_vs_7DMA_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(pos_rate_tr, case_7dma)) +
+  geom_point(color = "#00b0ba") +
+  stat_poly_line(color = "#00b0ba") +
+  stat_poly_eq(color = "black") +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Clinical Case Seven Day Moving Average") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_vs_7DMA_Y2.png"))
+```
+
+No significant change in data splits
+
+### Assay positivity vs tests administered
+
+```{r}
+#full dataset
+data_clean %>% 
+  ggplot(aes(pos_rate_tr,test_admin)) +
   geom_point() +
@@
   stat_poly_eq() +
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-  xlab(label = "Log10 Daily Viral Load") +
-  ylab(label = "Log10 Clinical Case Seven Day Moving Average") +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR tests administered") +
   theme_light()
 
-ggsave(here("figures/exploratory/covariate/total_copies_vs_7DMA.png"))
-```
+ggsave(here("figures/exploratory/covariate/assaypos_testadmin.png"))
 
-### Log10 viral load vs log10 daily positive cases
-
-```{r}
-all_data_combine %>% 
-  drop_na() %>% 
-  group_by(date, target, facility) %>% 
-  summarize(mean_daily_copies = mean(total_copies), 
-            mean_pcr_pos = mean(pcr_pos)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies = sum(mean_daily_copies),
-            mean_pcr_pos = mean_pcr_pos) %>% 
-  ggplot(aes(log10(sum_mean_copies), log10(mean_pcr_pos), color = target)) +
+#data split before FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(pos_rate_tr,test_admin)) +
   geom_point() +
@@
   stat_poly_eq() +
-  scale_color_manual(name = "Target",values = c("#F67280", "#355C7D")) +
-  xlab(label = "Log10 Daily Viral Load") +
-  ylab(label = "Log10 Daily Positive Clinical Cases") +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR tests administered") +
   theme_light()
 
-ggsave(here("figures/exploratory/covariate/total_copies_vs_pcr_pos.png"))
+ggsave(here("figures/exploratory/covariate/assaypos_testadmin_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(pos_rate_tr,test_admin)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR tests administered") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_testadmin_Y2.png"))
 ```
 
-### Log10 viral load vs log10 daily positive cases
+RSQ is low across all data sets, but highest in earlier dates than in full data or later dates
+
+### Assay positivity vs percent of positive tests
 
 ```{r}
-all_data_combine %>% 
-  drop_na() %>% 
-  group_by(date, facility) %>% 
-  summarize(mean_daily_copies = mean(total_copies), 
-            mean_pcr_pos = mean(pcr_pos)) %>% 
-  ungroup(facility) %>% 
-  summarize(sum_mean_copies = sum(mean_daily_copies),
-            mean_pcr_pos = mean_pcr_pos) %>% 
-  ggplot(aes(log10(sum_mean_copies), log10(mean_pcr_pos))) +
-  geom_point(color = "#80256D") +
-  stat_poly_line(color = "#80256D") +
-  stat_poly_eq(color = "#80256D") +
-  xlab(label = "Log10 Daily Viral Load") +
-  ylab(label = "Log10 Daily Reported Clinical Cases") +
+#full dataset
+data_clean %>% 
+  ggplot(aes(pos_rate_tr,percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR test positivity") +
   theme_light()
 
-ggsave(here("figures/exploratory/covariate/total_copies_avg_vs_pcr_pos.png"))
+ggsave(here("figures/exploratory/covariate/assaypos_testpos.png"))
+
+#data split before FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(pos_rate_tr,percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR test positivity") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_testpos_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(pos_rate_tr,percent_pos_test)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "PCR test positivity") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_testpos_Y2.png"))
 ```
 
-### Sample percent positive vs log10 7DMA
+There is a slight decrease in RSQ in the later dates than in the full set or the earlier dates.
+
+### Assay positivity vs daily reported cases
 
 ```{r}
+#full dataset
+data_clean %>% 
+  ggplot(aes(pos_rate_tr,pcr_pos)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Daily reported cases") +
+  theme_light()
 
+ggsave(here("figures/exploratory/covariate/assaypos_dailycases.png"))
+
+#data split before FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) %>%
+  ggplot(aes(pos_rate_tr,pcr_pos)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Daily reported cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_dailycases_Y1.png"))
+
+#data split after FEB 28 2022
+data_clean %>% 
+  filter(between(date, as.Date("2022-02-28"), as.Date("2022-12-21"))) %>%
+  ggplot(aes(pos_rate_tr,pcr_pos)) +
+  geom_point() +
+  stat_poly_line() +
+  stat_poly_eq() +
+  xlab(label = "qPCR assay positivity") +
+  ylab(label = "Daily reported cases") +
+  theme_light()
+
+ggsave(here("figures/exploratory/covariate/assaypos_dailycases_Y2.png"))
 ```
+
+The highest RSQ value was in the earlier dates than in the full set or later dates

### 2023-04-24 code/modeling_code/1_data_splitting.Rmd
fixing issues

--- 2023-03-17 code/modeling_code/data_splitting.Rmd
@@
 
-```{r}
+```{r, message=FALSE}
 library(tidyverse) 
@@
 ```{r}
-covid_ww_data <- readRDS(here("data/processed_data/final_combined_data.RDS"))
+covid_ww_data <- readRDS(here("data/processed_data/final_sum_combined_data.RDS"))
 covid_ww_data <- covid_ww_data %>% drop_na()
@@
 
-Since our data set is large enough, we can split it into a training set and a testing set.
-
-First though, I want to split the data such that one set contains observations prior to July 2022 and the latter contains observations after July 2022.
+I want to split the data such that one set contains observations prior to February 28, 2022 and the latter contains observations after February 28, 2022.
 From my exploratory analysis, namely, comparing temporal trends of wastewater viral load with clinical case moving averages, I noted that there appears to be a disconnect in trends after the Omicron BA.5 wave (Late summer/fall 2022).
-As such, I would like to build models using data prior to this time and then test the model on the latter portion of the data to see if there is a difference.
+As such, I would like to build models using data prior to this time and then test the model on the latter portion of the data.
 
@@
 ```{r}
-set.seed(420)
-train_dates <- covid_ww_data %>% filter(collection_num <= 193)
+train_dates <- covid_ww_data %>% filter(between(date, as.Date("2020-06-30"), as.Date("2022-02-28"))) # n=151
 
-test_dates <- covid_ww_data %>% filter(collection_num > 193)
+test_dates <- covid_ww_data %>% filter(between(date, as.Date("2022-03-02"), as.Date("2022-12-21"))) # n=73
 
 saveRDS(test_dates, here("data/modeling_data/test_dates.RDS"))
+
 ```
@@
 ```{r}
+set.seed(420)
 data_split <- initial_split(train_dates, prop = 3/4)
 
-train_data <- training(data_split)
-test_data <- testing(data_split)
+train_data <- training(data_split) # n=113
+test_data <- testing(data_split) # n=38
 

### 2023-04-24 code/modeling_code/PPBoth_7DMA_model.Rmd
fixing issues

---
title: "PP_7DMA_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Percent positive technical replicates per sample

Outcome: Clinical case seven day moving average

## Load packages and data

```{r, message=FALSE}
library(tidyverse)
library(tidymodels)
library(here)
library(ggpmisc)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data_both.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data_both.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set PP as predictor and 7DMA as outcome

```{r}
recipe_7dma <- recipe(pcr_pos_7dma ~ percent_pos, data = train_data)
```

## Create workflow

```{r}
workflow_7dma <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_7dma)
```

## Create model using workflow

```{r}
set.seed(828)
fit_7dma <- workflow_7dma %>% 
  fit(data = train_data)

tidy(fit_7dma)
```

## Use model on the training data

```{r}
aug_test <- augment(fit_7dma, train_data)
aug_test %>% select(pcr_pos_7dma, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
huber <- aug_test %>% huber_loss(truth = pcr_pos_7dma, .pred)
PP_7DMA_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_metrics, here("data/modeling_data/metrics/PP_7DMA_performance.RDS"))

PP_7DMA_metrics

```

## Plot actual value vs predicted value

```{r, message=FALSE}
aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
  geom_point() +
  stat_p
  stat
ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
```

## Plot residual values vs predicted values

```{r, message=FALSE}
aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point() +
  stat_poly_line()

ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
```

## Plot residual values vs actual values

```{r}
aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
aug_residual %>% ggplot(aes(pcr_pos_7dma, residual)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq() 
```

## Plot 7DMA and prediction temporally

```{r}
aug_test %>% ggplot(aes(x=date)) +
  geom_point(aes(y=pcr_pos_7dma), color = "blue") +
  geom_line(aes(y=pcr_pos_7dma), color = "blue") +
  geom_point(aes(y=.pred), color = "green") +
  geom_line(aes(y=.pred), color = "green")

```

## Use model on testing data

```{r}
aug_test2 <- augment(fit_7dma, test_data)
aug_test2 %>% select(pcr_pos_7dma, .pred)
```

```{r}
aug_test2 %>% ggplot(aes(x=date)) +
  geom_point(aes(y=pcr_pos_7dma), color = "blue") +
  geom_line(aes(y=pcr_pos_7dma), color = "blue") +
 geom_point(aes(y=.pred), color = "green") +
  geom_line(aes(y=.pred), color = "green")
```

## Evaluate model performance on testing set

```{r}
rmse <- aug_test2 %>% rmse(truth = pcr_pos_7dma, .pred)
rsq <- aug_test2 %>% rsq(truth = pcr_pos_7dma, .pred)
huber <- aug_test2 %>% huber_loss(truth = pcr_pos_7dma, .pred)
PP_7DMA_test_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_test_metrics, here("data/modeling_data/metrics/PP_7DMA_test_performance.RDS"))

PP_7DMA_test_metrics
```

## Use model on testing dates

```{r}
aug_test3 <- augment(fit_7dma, test_dates_both)
aug_test3 %>% select(pcr_pos_7dma, .pred)
```

## Evaluate model performance on testing dates

```{r}
rmse <- aug_test3 %>% rmse(truth = pcr_pos_7dma, .pred)
rsq <- aug_test3 %>% rsq(truth = pcr_pos_7dma, .pred)
huber <- aug_test3 %>% huber_loss(truth = pcr_pos_7dma, .pred)
PP_7DMA_BA5_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_BA5_metrics, here("data/modeling_data/metrics/PP_7DMA_BA5_performance.RDS"))

PP_7DMA_BA5_metrics
```

```{r}
aug_test3 %>% ggplot(aes(x=date)) +
  geom_line(aes(y=pcr_pos_7dma), color = "#621940", size = 2) +
  theme_light()

```

```{r, message=FALSE}
 aug_test3 %>% ggplot(aes(x=date)) +
   geom_line(aes(y=pcr_pos_7dma), color = "#1c246c", size = 2) +
   geom_line(aes(y=.pred), color = "#ec4998", size = 2) +
   theme_light() +
     ylab(label = "Predicted 7-Day Moving Average") +
  xlab(label = "Date") +
  theme_light()

ggsave(here("figures/modeling/predicted_7DMA.png"))
```

### 2023-04-24 code/processing_code/percent_pos.Rmd
fixing issues

---
title: "Calculating percent positive assays"
author: "Leah Lariscy"
date: "2023-03-26"
output: html_document
---

## Load packages

```{r}
library(tidyverse)
library(here)
```

## Load data

```{r}
wbe <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
```

## Calculate percent positive qPCR assays

```{r}
wbe_pos <- wbe %>% group_by(date, target, facility, biological_replicate) %>% 
  #n positive per biological rep
  summarise(n=n(),
            n.miss=sum(is.na(ct)),
            n.pos=n-n.miss) %>% 
  #n positive per target/facility/date
  ungroup(biological_replicate) %>% 
  summarise(n=sum(n),
            n.miss=sum(n.miss),
            n.pos=n-n.miss) %>% 
  #n positive per target/date
  ungroup(facility) %>% 
  summarise(n=sum(n),
            n.miss=sum(n.miss),
            n.pos=n-n.miss) %>%
  ungroup()

wbe_pos
```

### Separate N1 and N2

```{r}
wbe_pos_n1 <- wbe_pos %>% filter(target == "N1")

wbe_pos_n2 <- wbe_pos %>% filter(target == "N2")

saveRDS(wbe_pos_n1, here("data/processed_data/percent_pos_n1.rds"))
saveRDS(wbe_pos_n2, here("data/processed_data/percent_pos_n2.rds"))
```

### Sum N1 and N2

```{r}
# n positive per date
wbe_pos_both <- wbe_pos %>% group_by(date) %>% 
   summarise(n=sum(n),
            n.miss=sum(n.miss),
            n.pos=n-n.miss) %>% 
  mutate(percent_pos = (n.pos/n)*100) 
wbe_pos_both
```

### Avg N1 and N2

```{r}
wbe_pos_avg <- wbe_pos %>% group_by(date) %>% 
   summarise(n=sum(n),
            n.miss=sum(n.miss),
            n.pos=n-n.miss) %>% 
  mutate(percent_pos = (n.pos/n)*100) 
wbe_pos_avg
```

## Load COVID data and combine with qPCR data

```{r}
covid <- readRDS(here("data/processed_data/all_covid_combined.rds"))

covid_n1 <- full_join(wbe_pos_n1, covid, by = "date") %>% drop_na(target)

covid_n2 <- full_join(wbe_pos_n2, covid, by = "date") %>% drop_na(target)

covid_both <- full_join(wbe_pos_both, covid, by = "date") %>% drop_na(n)

saveRDS(covid_n1, here("data/processed_data/covid_ww_n1.rds"))
saveRDS(covid_n2, here("data/processed_data/covid_ww_n2.rds"))
saveRDS(covid_both, here("data/processed_data/covid_ww_both.rds"))
```

```{r}
covid_both %>% ggplot(aes(date, percent_pos)) +
geom_line(size =1.5, color = '#634499') +
  xlab(label = "Date") +
  ylab(label = "Percent Positive qPCR assays") +
  theme_light()

ggsave(here("figures/exploratory/covariate/percent_pos_time_plot.png"))
```

```{r}
  covid_both %>% 
    filter(between(date, as.Date("2020-06-30"), as.Date("2022-12-21"))) %>% 
    ggplot(aes(percent_pos, pcr_pos_7dma)) +
  geom_point(color = "#634499") +
  stat_poly_line(color = "#634499") +
  stat_poly_eq(color = "black") +
  xlab(label = "Percent Positve N1 & N2 Assays") +
  ylab(label = "Clinical Case Seven Day Moving Average") +
  theme_light()

  ggsave(here("figures/exploratory/covariate/PP_vs_7DMA.png"))
```

```{r}
 covid_both %>% 
    filter(between(date, as.Date("2020-06-30"), as.Date("2021-06-30"))) %>% 
    ggplot(aes(percent_pos, pcr_pos)) +
  geom_point(color = "#4dd091") +
  stat_poly_line(color = "#4dd091") +
  stat_poly_eq(color = "black") +
  xlab(label = "Percent Positve N1 & N2 Assays") +
  ylab(label = "Daily Reported Clinical Cases") +
  theme_light()

 ggsave(here("figures/exploratory/covariate/PP_daily_cases.png"))
```

### 2023-04-24 code/processing_code/processing.qmd
fixing issues

@@
 
-recovery_output \<- read_csv(here("data/raw_data/recovery_data.csv"))
-
-recovery_input \<- read_csv(here("data/raw_data/calfguard.csv"))
-
-```{r}
+```{r, message=FALSE}
 here()
 # Load CFX N1 data
-n1 <- read_csv(here("data/raw_data/cfx_n1.csv"))
+n1_1 <- read_csv(here("data/raw_data/n1_data_year1.csv")) #year 1 data
+n1 <- read_csv(here("data/raw_data/cfx_n1.csv")) #year 2 data
 
 # Load CFX N2 data
-n2 <- read_csv(here("data/raw_data/cfx_n2.csv"))
+n2_1 <- read_csv(here("data/raw_data/n2_data_year1.csv")) #year 1 data
+n2 <- read_csv(here("data/raw_data/cfx_n2.csv")) #year 2 data
 
@@
 
-### Catchment-level COVID-19 Case Data
-
-`covid_case_wrf` \<- Count of confirmed cases of COVID-19 on a given day from each WRF catchment zone in ACC.
-
-```{r}
-#I don't have access to this data yet, waiting on DPH partner to send.
-```
-
 ## Data cleaning and wrangling
 
-Currently, I have 7 unique sets of data. In order to do analysis on the data down the road, I want to figure out how to best combine these data sets.
+Currently, I have 6 datasets. In order to do analysis on these data down the road, I want to figure out how to best combine them.
 
@@
 2.  CFX N2 results
-3.  Plant flow and total suspended solids
+3.  WWTP flow rate and total solids
 4.  County-level COVID symptoms
@@
 6.  County-level COVID tests administered
-7.  Catchment-level COVID confirmed cases
-8.  CFX recovery input
-9.  CFX recovery output
 
@@
 
+### Remove unnecessary variables
+
+```{r}
+#N1
+n1_1 <- n1_1 %>% select(c(sample_date, collection_num, sample_id, target, ct))
+n1 <- n1 %>% select(c(sample_date, collection_num, sample_id, target, ct)) %>% 
+  filter(between(sample_date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+
+#N2
+n2_1 <- n2_1 %>% select(c(sample_date, collection_num, sample_id, target, ct))
+n2 <- n2 %>% select(c(sample_date, collection_num, sample_id, target, ct)) %>% 
+  filter(between(sample_date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+```
+
 ### Calculate SARS-CoV-2 concentrations from CFX data for both N1 and N2
@@
 
-copy_num_uL_rxn = as.numeric(10\^((ct-36.046)/-3.5293)), #use standard curve slope here
+Year1:
+
+Year 2: copy_num_uL_rxn = as.numeric(10\^((ct-36.046)/-3.5293)), #use standard curve slope here
 
@@
 ```{r}
-#N1 data here
+#year 1 N1 data here
+n1_1 <- n1_1 %>% mutate(ct = as.numeric(ct), 
+                copy_num_uL_rxn = as.numeric(10^((ct-34.008)/-3.389)),
+                copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))
+
+#year 2 N1 data here
 n1 <- n1 %>% mutate(ct = as.numeric(ct), 
               copy_num_uL_rxn = as.numeric(10^((ct-36.046)/-3.5293)), #use standard curve slope here
-              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004), #replace NAs with limit of detection (0.004 copies per uL of the reaction)
-              run_date = as.Date(run_date))
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004)) #replace NAs with limit of detection (0.004 copies per uL of the reaction)
 
-#N2 data here (because they have different standard curve equations)
+#N2 year 1 data here
+n2_1 <-n2_1 %>% mutate(ct = as.numeric(ct),
+                copy_num_uL_rxn = as.numeric(10^((ct-32.416)/-3.3084)),
+                copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))
+
+#N2 year 2 data here
 n2 <- n2 %>% mutate(ct = as.numeric(ct),
               copy_num_uL_rxn = as.numeric(10^((ct-37.731)/-3.2505)),
-              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004),
-              run_date = as.Date(run_date))
+              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))
 
+```
+
+### Combining year 1 and 2 data
+
+```{r}
+#N1
+n1 <- bind_rows(n1_1, n1)
+tibble(n1)
+
+#N2
+n2 <- bind_rows(n2_1, n2)
+tibble(n2)
 ```
@@
          facility = ordered(facility, levels = c("A", "B", "C")), #set facility to factor and then set factor levels
-         copy_num_L = copy_num_uL_rxn*20/5*60/280*1000*1000) %>% 
-  filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+         copy_num_L = copy_num_uL_rxn*20/5*60/280*1000*1000)
 
@@
 ```{r}
-n1_n2_plant <- left_join(wbe_year2_avg, plant_clean, by = c("date", "facility")) %>% 
-  mutate(total_copies = avg_copy_L * influent_flow_L)
+n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date", "facility")) %>% 
+  mutate(total_copies = copy_num_L * influent_flow_L)
 
@@
 all_data <- full_join(n1_n2_plant, covid, by = "date") %>% 
-   filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))
+   filter(between(date, as.Date("2020-06-30") , as.Date("2022-12-21")))
 
@@
 
-### Combining recovery data sets
-
-```{r}
-recovery_calc <- left_join(recovery_output_clean, recovery_input_clean, by = "cg_num") %>% 
-  mutate(percent_recovery = (copies_ul_sample/copies_ul_input)*100) %>% 
-  select(facility, collection_num, percent_recovery)
-
-saveRDS(recovery_calc, here("data/processed_data/recovery_percentages.rds"))
-```
-
 ### Calculate percent positive of technical reps
@@
 
+#### Percent positive technical reps for N1 only
+
+```{r}
+wbe_tr_n1 <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
+  filter(target == "N1") %>% 
+  dplyr::summarise(
+    n=n(), 
+    n.miss=sum(is.na(ct)), 
+    ct.mean=mean(ct,na.rm=T), 
+    ct.sd=sd(ct,na.rm=T)
+  ) %>% 
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(n.pos = n - n.miss) %>% 
+  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
+  mutate(percent_pos = (n.pos/n)*100)
+wbe_tr_n1
+```
+
+#### Percent positive N2 only
+
+```{r}
+wbe_tr_n2 <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
+  filter(target == "N2") %>% 
+  dplyr::summarise(
+    n=n(), 
+    n.miss=sum(is.na(ct)), 
+    ct.mean=mean(ct,na.rm=T), 
+    ct.sd=sd(ct,na.rm=T)
+  ) %>% 
+  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup() %>% 
+  mutate(n.pos = n - n.miss) %>% 
+  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
+  mutate(percent_pos = (n.pos/n)*100)
+wbe_tr_n2
+```
+
 ### Calculate percent positive per biological replicate
 
-`pos_rate_tr` will calculate the percento of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)
+`pos_rate_tr` will calculate the percent of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)
 
@@
     bio.ct.sd = sd(ct.mean, na.rm=T), 
-    pos_rate_tr = mean(percent_pos),
-    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
-  ) %>%
+    pos_rate_tr = mean(percent_pos)) %>%
   mutate(percent_pos_bio = (n.bio.non.miss/n.bio)*100) %>% 
@@
 
+### Calculate percent positive per collection date
+
+```{r}
+wbe_coll <- wbe_samp %>%
+  group_by(date) %>% 
+  dplyr::summarise(
+    n.pos = sum(!is.na(n.pos)),
+    n.miss = sum(is.na(n.miss)),
+    pos_rate_tr = mean(pos_rate_tr)) %>%
+   mutate(prop.pos = n.pos/(n.pos+n.miss)*100) %>% 
+  #mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
+  ungroup()
+```
+
 ### Combine all data: wastewater, clinical cases, percent positive
@@
 
-final_data
 ```
@@
             percent_pos_test = mean(percent_pos_test),
-            case_7dma = mean(pcr_pos_7dma)) %>% 
+            case_7dma = mean(pcr_pos_7dma),
+            pcr_pos = mean(pcr_pos),
+            test_admin = mean(pcr_tests)) %>% 
   ungroup(target) %>% 
@@
             percent_pos_test = mean(percent_pos_test),
-            case_7dma = mean(case_7dma)) %>% 
+            case_7dma = mean(case_7dma),
+            pcr_pos = mean(pcr_pos),
+            test_admin = mean(test_admin)) %>% 
   ungroup() %>% 
-  mutate(log10_VL = log10(avg_VL))
+  mutate(log10_VL = log10(avg_VL)) %>% 
+  drop_na()
+
+table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
+table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
+table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
+table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
+table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"
 
 saveRDS(final_data_sum, here("data/processed_data/final_sum_combined_data.RDS"))
+
+final_data_sum
+
 ```
+
+```{r}
+
+table1 <- (table1(~ log10_VL + pos_rate_tr + case_7dma + pcr_pos + percent_pos_test, data=final_data_sum)) 
+saveRDS(table1, here("results/manuscript_table1.rds"))
+```

### 2023-04-24 products/manuscript/Manuscript.qmd
fixing issues

--- 2023-03-18 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Data Analysis Project"
+title: "Wastewater Surveillance for SARS-CoV-2 Data Analysis Project"
 subtitle: ""
@@
 format:
-  docx:
+  html:
     toc: false
     number-sections: true
+    number-depth: 2
     highlight-style: github
-bibliography: ../dataanalysis_template_references.bib
+bibliography: ../citation.bib
 csl: ../apa.csl
@@
 ```{r, echo=FALSE, message=FALSE}
-# load a few R packages
+# load packages
 library(here)
 library(knitr)
+library(tidyverse)
+library(skimr)
 ```
 
-# Summary/Abstract
+# Abstract
 
-*Write a summary of your project.*
+Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.
+Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.
+Wastewater-based detection methods have been utilized across the globe as an independent and parallel indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends.
+WBE methods may also become a primary tool in understanding transmission dynamics as case under-reporting increases.
 
@@
 
-Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.
+Paragraph 1: Impact of COVID-19 on public health, clinical case under-reporting
 
-Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.
+Paragraph 2: Wastewater surveillance is possible because of excretory functions and stability of viral particles in wastewater, surveillance prevalence and correlations with clinical case data
 
-Wastewater-based detection methods have been utilized across the globe as an independent indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends \[4\].
-WBE methods may also become a primary tool in understanding transmission dynamics as case under reporting increases.
+Paragraph 3: Surveillance program background. Aims, County-level parallel analysis of wastewater surveillance data and reported clinical cases
 
@@
 
-24-hour composite samples collected twice weekly, stored at 4C until ready for extraction. 6 extraction replicates were performed for each collection date at treatment facility.
+### SARS-CoV-2 wastewater surveillance data
 
-2019-nCoV N1 and N2 primer-probe sets were used for universal detection of SARS-CoV-2.
+Sample collection for wastewater surveillance began June 30, 2020 and occurred twice weekly (excluding major holidays) through December 21, 2022.
+Raw wastewater influent was collected from three Wastewater Treatment Plants in Athens-Clarke County and viral RNA was extracted.
+RT-qPCR was used to calculate the cycle threshold (Ct) value of each reaction, which is used to estimate SARS-CoV-2 copies per liter of wastewater in the data cleaning process (Section 3.2).
 
-The concentration per reaction of each gene target was determined by the corresponding standard curve, then sample concentration in copies per liter was determined by Equation 1.
-Total daily copies per day per WRF was determined by Equation 2.
+### Wastewater treatment plant data
 
-## Questions/Hypotheses to be addressed
+Wastewater influent flow data, including millions of gallons per day (MGD) and total suspended solids (TSS), were collected for each corresponding sampling date.
 
-Is there a relationship between viral load and clinical case observations?
+### COVID-19 clinical data
 
-Is there a relationship between qPCR percent positive and clinical case observations?
+COVID-19 clinical data were obtained from the Georgia Department of Public Health website [@COVID19StatusReport] which includes COVID-19 reported clinical case data for the state of Georgia.
+These data include reported cases, demographics, comorbidities, deaths, and more.
+However, not all data is necessary for analysis and will be subset to include only symptom onset, positive case, and administered test data for Clarke County.
+(Section 3.2).
 
-What do viral load trends look like over time?
+## Hypotheses to be addressed
 
-What do clinical case trends look like over time?
-
-How do each of these temporal trends compare to one another over time?
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
+This study aims to conduct a parallel analysis between SARS-CoV-2 wastewater surveillance data and reported COVID-19 clinical case data in Athens-Clarke County, Georgia.
+It is expected that wastewater quantifications and reported case data will have a positive linear relationship.
+It is also hypothesized that certain wastewater-clinical relationships will be stronger than others, see Table 1 for variables of interest.
 
@@
 
-## Data aquisition
+## Data acquisition
 
-Wastewater SARS-CoV-2 data was obtained through weekly surveillance efforts utilizing RT-qPCR methods.
+### Wastewater surveillance data
 
-Water Reclamation Facility data was obtained through data share directly from the WRF representatives.
+24-hour composite raw wastewater influent samples were collected twice-weekly from three treatment facilities in Clarke County and stored at 4°C until ready for extraction.
+280 𝜇L of wastewater in replicates (n=6) was aliquoted and direct RNA extraction (Zymo Research, R2042) was performed on the same day as sample collection.Each extraction replicate was reverse transcribed and amplified using RT-qPCR and a SARS-CoV-2 specific primer-probe panel (IDT, 10006713), done in technical replicates (n=3).
+As part of the qPCR panel, two gene targets on the N-gene of the SARS-CoV-2 genome were used for amplification.
+The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus \[\@\].
+Cycle threshold (Ct) values were then calculated for each PCR reaction, which were then used to estimate SARS-CoV-2 viral load (See Section 3.2 for more detail).
 
-COVID data downloaded from Georgia DPH website.
+### Wastewater treatment plant data
+
+For each treatment plant and collection date, MGD and TSS was shared by plant operators directly with our team and compiled throughout the study period.
+
+### COVID-19 clinical data
+
+Reported case and testing data were downloaded from Georgia Department of Public Health's COVID-19 Status Report Page in January 2023.
 
@@
 
-*qPCR data*
+### Calculating wastewater viral load
 
-WWTP data
+Amplification standard curve equations where generated from serially diluted synthetic controls (Supplementary Table).
+This was done for N1 and N2 gene targets for both surveillance years because RT-qPCR methods were altered after the first year.
+Viral copies per microliter (cp/uL) of each PCR reaction was estimated by transforming the Ct value using the appropriate standard curve equation.
+For reactions with no detection, concentration was set to the theoretical limit of detection (0.004 cp/uL).
+Data sets for each standard curve was transformed separately before being combined.
+After combining all qPCR data, copies per liter (cp/L) of wastewater was calculated for each technical replicate using Equation 1.
+Average cp/L of wastewater was determined by taking the mean of all technical replicates per sample per gene target.
 
-*DPH COVID data*
+INSERT EQUATION 1
+
+WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day.
+qPCR data was then combined with WWTP data and total viral load per sample was calculated by multiplying average copies per liter by the flow rate.
+Average viral load for each sample was then determined by taking the mean of the viral load for both N1 and N2 gene targets.
+County-level daily viral load was calculated by taking the sum of the average viral load from each of the three WWTP samples.
+
+### Calculating wastewater qPCR assay positivity
+
+qPCR assay positivity was determined for each collection date by finding the proportion of assays that detected SARS-CoV-2 out of the total number of assays performed.
+This includes all biological and technical replicate assays that were performed for a given sampling date.
+County-level assay positivity data was then combined with county-level viral load data.
+
+### Adding COVID-19 clinical data
+
+Of the downloaded Georgia DPH data, the data set containing positive case counts was retained.
+Of these data, daily reported case counts and case seven-day moving average were kept for analysis.
+These data were then subset to include only data from Clarke County collected during the surveillance period and combined with wastewater data.
 
@@
 
-Simple analyses: Wastewater and clinical data over time, linear regression model of wastewater vs clinical case data
+### Feature selection
 
-Full analyses: Wastewater data as predictor for clinical case outcomes
+Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where
 
-Steps I took to build and tune model
+### Data splitting
+
+Compare relationship differences between year 1 and year 2
+
+### Model training and fitting
+
+Trained linear regression models on year 1 data using cross validation
+
+Performance validation, metrics?
+
+### Model testing
+
+The trained model was then tested on the year 2 data
 
@@
 
-## Exploratory/Descriptive analysis
+```{r, echo=FALSE, message=FALSE}
+
+readRDS(here("results/manuscript_table1.rds"))
+```
 
@@
 
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
 {{< pagebreak >}}

### 2023-04-25 code/modeling_code/2_log10VL_7DMA_model.Rmd
update manuscript

--- 2023-03-20 code/modeling_code/log10VL_7DMA_model.Rmd
@@
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL, data = train_data)
+recipe_7dma <- recipe(case_7dma ~ log10_VL, data = train_data)
 ```
@@
 aug_test <- augment(fit_7dma, train_data)
-aug_test %>% select(pcr_pos_7dma, .pred)
+aug_test %>% select(case_7dma, .pred)
 ```
@@
 ```{r}
-rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
-rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
 log10VL_7DMA_metrics <- full_join(rmse, rsq)
@@
 ```{r}
-aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
+aug_test %>% ggplot(aes(case_7dma, .pred)) +
   geom_point() +
@@
 ```{r}
-aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
+aug_residual <- aug_test %>% mutate(residual = case_7dma - .pred)
 aug_residual %>% ggplot(aes(.pred, residual)) +

### 2023-04-25 code/modeling_code/3_log10VL_TestPos_model.Rmd
update manuscript

---
title: "log10VL_testpos_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: log10 wastewater viral load

Outcome: Clinical case seven day moving average

## Load packages and data

```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
```

## Define the model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg()
```

## Create recipe using training data

This is where we will set log10VL as predictor and testpos as outcome

```{r}
recipe_testpos <- recipe(percent_pos_test ~ log10_VL, data = train_data)
```

## Create workflow

```{r}
workflow_testpos <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_testpos)
```

## Create model using workflow

```{r}
set.seed(626)
fit_testpos <- workflow_testpos %>% 
  fit(data = train_data)

tidy(fit_testpos)
```

## Use model on the training data

```{r}
aug_test <- augment(fit_testpos, train_data)
aug_test %>% select(percent_pos_test, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
log10VL_testpos_metrics <- full_join(rmse, rsq)

saveRDS(log10VL_testpos_metrics, here("data/modeling_data/metrics/log10VL_testpos_performance.RDS"))

log10VL_testpos_metrics

```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(percent_pos_test, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/log10VL_testpos_predict.png"))
```

## Plot residual values

```{r}
aug_residual <- aug_test %>% mutate(residual = percent_pos_test - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

ggsave(here("data/modeling_data/metrics/log10VL_testpos_residual.png"))
```

## Final fit on test data, then on test dates

```{r}
#testing data
fit_test_data <- workflow_testpos %>% 
  fit(data = test_data)

aug_test <- augment(fit_test_data, test_data)
aug_test %>% select(percent_pos_test, .pred)

rmse <- aug_test %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
test_metrics <- full_join(rmse, rsq)

test_metrics
```

```{r}
#testing dates
fit_test_dates <- workflow_testpos %>% 
  fit(data = test_dates)

aug_test_dates <- augment(fit_test_dates, test_dates)
aug_test_dates %>% select(percent_pos_test, .pred)

rmse <- aug_test_dates %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test_dates %>% rsq(truth = percent_pos_test, .pred)
test_dates_metrics <- full_join(rmse, rsq)

test_dates_metrics
```

## Plot testing dates predictions vs truth

```{r}
aug_test_dates %>% ggplot(aes(x=date)) +
  stat_smooth(aes(y=percent_pos_test), method = "loess", span = 0.5, color = "black") +
  stat_smooth(aes(y=.pred), method = "loess", span = 0.5)
```

### 2023-04-25 code/modeling_code/4_AssayPos_7DMA_model.Rmd
update manuscript

--- 2023-03-20 code/modeling_code/PP_7DMA_model.Rmd
@@
 ---
-title: "PP_7DMA_model"
+title: "assaypos_7dma_model"
 author: "Leah Lariscy"
@@
 
-Predictor: Percent positive technical replicates per sample
+Predictor: qPCR assay positivity
 
@@
 
-This is where we will set PP as predictor and 7DMA as outcome
+This is where we will set assaypos as predictor and 7dma as outcome
 
@@
 rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
-PP_7DMA_metrics <- full_join(rmse, rsq)
+assaypos_7dma_metrics <- full_join(rmse, rsq)
 
-saveRDS(PP_7DMA_metrics, here("data/modeling_data/metrics/PP_7DMA_performance.RDS"))
+saveRDS(assaypos_7dma_metrics, here("data/modeling_data/metrics/assaypos_7dma_performance.RDS"))
 
-PP_7DMA_metrics
+assaypos_7dma_metrics
 
@@
   stat_smooth(method = "lm")
-ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_7dma_predict.png"))
 ```
@@
 
-ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_7dma_residual.png"))
 ```
+
+## Final fit on test data, then on test dates
+
+```{r}
+#testing data
+fit_test_data <- workflow_7dma %>% 
+  fit(data = test_data)
+
+aug_test <- augment(fit_test_data, test_data)
+aug_test %>% select(case_7dma, .pred)
+
+rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
+test_metrics <- full_join(rmse, rsq)
+
+test_metrics
+```
+
+```{r}
+#testing dates
+fit_test_dates <- workflow_7dma %>% 
+  fit(data = test_dates)
+
+aug_test_dates <- augment(fit_test_dates, test_dates)
+aug_test_dates %>% select(case_7dma, .pred)
+
+rmse <- aug_test_dates %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test_dates %>% rsq(truth = case_7dma, .pred)
+test_dates_metrics <- full_join(rmse, rsq)
+
+test_dates_metrics
+```
+
+## Plot testing dates predictions vs truth
+
+```{r}
+aug_test_dates %>% ggplot(aes(x=date)) +
+  stat_smooth(aes(y=case_7dma), method = "loess", span = 0.5, color = "black") +
+  stat_smooth(aes(y=.pred), method = "loess", span = 0.5)
+```

### 2023-04-25 code/modeling_code/5_AssayPos_TestPos_model.Rmd
update manuscript

--- 2023-04-25 code/modeling_code/3_log10VL_TestPos_model.Rmd
@@
 ---
-title: "log10VL_testpos_model"
+title: "assaypos_7dma_model"
 author: "Leah Lariscy"
@@
 
-Predictor: log10 wastewater viral load
+Predictor: qPCR assay positivity
 
-Outcome: Clinical case seven day moving average
+Outcome: PCR test positivity rate (daily)
 
@@
 
-This is where we will set log10VL as predictor and testpos as outcome
+This is where we will set assaypos as predictor and 7dma as outcome
 
 ```{r}
-recipe_testpos <- recipe(percent_pos_test ~ log10_VL, data = train_data)
+recipe_7dma <- recipe(percent_pos_test ~ pos_rate_tr, data = train_data)
 ```
@@
 ```{r}
-workflow_testpos <- workflow() %>% 
+workflow_7dma <- workflow() %>% 
   add_model(lm_mod) %>% 
-  add_recipe(recipe_testpos)
+  add_recipe(recipe_7dma)
 ```
@@
 set.seed(626)
-fit_testpos <- workflow_testpos %>% 
+fit_7dma <- workflow_7dma %>% 
   fit(data = train_data)
 
-tidy(fit_testpos)
+tidy(fit_7dma)
 ```
@@
 ```{r}
-aug_test <- augment(fit_testpos, train_data)
+aug_test <- augment(fit_7dma, train_data)
 aug_test %>% select(percent_pos_test, .pred)
@@
 rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
-log10VL_testpos_metrics <- full_join(rmse, rsq)
+assaypos_7dma_metrics <- full_join(rmse, rsq)
 
-saveRDS(log10VL_testpos_metrics, here("data/modeling_data/metrics/log10VL_testpos_performance.RDS"))
+saveRDS(assaypos_7dma_metrics, here("data/modeling_data/metrics/assaypos_7dma_performance.RDS"))
 
-log10VL_testpos_metrics
+assaypos_7dma_metrics
 
@@
   stat_smooth(method = "lm")
-ggsave(here("data/modeling_data/metrics/log10VL_testpos_predict.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_7dma_predict.png"))
 ```
@@
 
-ggsave(here("data/modeling_data/metrics/log10VL_testpos_residual.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_7dma_residual.png"))
 ```
@@
 #testing data
-fit_test_data <- workflow_testpos %>% 
+fit_test_data <- workflow_7dma %>% 
   fit(data = test_data)
@@
 #testing dates
-fit_test_dates <- workflow_testpos %>% 
+fit_test_dates <- workflow_7dma %>% 
   fit(data = test_dates)

### 2023-04-25 code/modeling_code/dont_run/2_VL_7DMA_model.Rmd
update manuscript

---
title: "VL_7DMA_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

## Predictor and Outcome

Predictor: Log10 wastewater viral load

Outcome: Clinical case seven day moving average

## Load packages and data

```{r}
library(here)
library(tidyverse)
library(rsample) #Data split
library(tidymodels)
library(rpart) #Model fit
library(ranger) #Model fit
library(glmnet) #Model fit
library(rpart.plot)  #viz of decision tree
library(vip) #viz of variable importance plots
library(ggpmisc) #for adding linear regression to plots
```

```{r}
train_data <- readRDS(here("data/modeling_data/training_data.RDS")) %>% 
  as.data.frame()

test_data <- readRDS(here("data/modeling_data/testing_data.RDS")) %>% 
  as.data.frame()
```

## Define the linear regression model

The outcome is continuous, so we will use a linear regression model

```{r}
lm_mod <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression") %>% 
  translate()
```

## Define the null model

```{r}
null_mod <- null_model() %>% 
  set_engine("parsnip") %>% 
  set_mode("regression") %>%
  translate()
```

## Create recipe using training data

This is where we will set VL as predictor and 7DMA as outcome

```{r}
recipe_7dma <- recipe(case_7dma ~ log10_VL, data = train_data) %>% 
  step_nzv(all_nominal()) %>% 
  step_dummy(all_nominal())
```

## Create null model workflow

```{r}
workflow_null <- workflow() %>% 
  add_model(null_mod) %>% 
  add_recipe(recipe_7dma)
```

## Create training data workflow

```{r}
workflow_7dma <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_7dma)
```

## Create resampling, 5x5

```{r}
folds <- vfold_cv(train_data, v = 5, repeats = 5, strata = case_7dma)
```

## Fit model to training data and null model resampling

```{r}
set.seed(626)
fit_7dma_cv <- fit_resamples(workflow_7dma, resamples = folds)
fit_null_cv <- fit_resamples(workflow_null, resamples = folds)

collect_metrics(fit_7dma_cv)
collect_metrics(fit_null_cv)
```

## Fit model to training data

```{r}
fit(workflow_7dma, data = train_data)
```

## Use model on the training data

```{r}
aug_train <- augment(fit_7dma_cv, train_data)
aug_train %>% select(case_7dma, .pred)
```

## Use RMSE and RSQ to evaluate model performance

```{r}
rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
VL_7DMA_metrics <- full_join(rmse, rsq)

saveRDS(VL_7DMA_metrics, here("data/modeling_data/metrics/VL_7DMA_performance.RDS"))

VL_7DMA_metrics

```

## Plot actual value vs predicted value

```{r}
aug_test %>% ggplot(aes(case_7dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/VL_7DMA_predict.png"))
```

## Plot residual values

```{r}
aug_residual <- aug_test %>% mutate(residual = case_7dma - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

ggsave(here("data/modeling_data/metrics/VL_7DMA_residual.png"))
```

```{r}
aug_test %>% ggplot(aes(date)) +
  geom_line(aes(y=.pred)) +
  geom_line(aes(y=case_7dma), color = 'red')
```

## Use model on testing data

```{r}
aug_test2 <- augment(fit_7dma, test_data)
aug_test2 %>% select(case_7dma, .pred)

rmse <- aug_test2 %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test2 %>% rsq(truth = case_7dma, .pred)
huber <- aug_test2 %>% huber_loss(truth = case_7dma, .pred)
VL_7DMA_test_metrics <- bind_rows(rmse, rsq, huber)

VL_7DMA_test_metrics
```

## Use model on testing dates

```{r}
aug_test3 <- augment(fit_7dma, test_dates)
aug_test3 %>% select(case_7dma, .pred)

rmse <- aug_test3 %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test3 %>% rsq(truth = case_7dma, .pred)
huber <- aug_test3 %>% huber_loss(truth = case_7dma, .pred)
VL_7DMA_test_dates_metrics <- bind_rows(rmse, rsq, huber)

VL_7DMA_test_dates_metrics
```

```{r}
aug_test3 %>% ggplot(aes(x=date)) +
   geom_line(aes(y=case_7dma), color = "#1c246c", size = 2) +
   geom_line(aes(y=.pred), color = "#ec4998", size = 2) +
   theme_light() +
     ylab(label = "Predicted 7-Day Moving Average") +
  xlab(label = "Date") +
  theme_light()
```

### 2023-04-25 code/processing_code/processing.qmd
update manuscript

@@
 
-table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
-table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
-table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
-table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
-table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"
+#table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
+#table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
+#table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
+#table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
+#table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"
 

### 2023-04-25 products/manuscript/Manuscript.qmd
update manuscript

@@
 
-### Feature selection
-
-Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where
-
-### Data splitting
-
-Compare relationship differences between year 1 and year 2
-
-### Model training and fitting
-
-Trained linear regression models on year 1 data using cross validation
-
-Performance validation, metrics?
-
-### Model testing
-
-The trained model was then tested on the year 2 data
+Through exploration of time series and wastewater-clinical correlations, prominent relationships were identified for further analysis.
+Wastewater variables were used to train linear regression models to predict clinical observations.
+Model performance metrics (RSQ and RMSE) were assessed against each other and against a null model.
+Cross-validation was used to validate performances.
+The best model was then determined and used to predict values in a new set of data.
 
@@
 # Results
+
+## Time-series analysis
+
+## Correlation analysis
+
+Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where stronger wastewater-clinical correlations were observed between these variables than in others.
+These include wastewater viral load, wastewater assay positivity rate, clinical case seven-day moving average, and test administration positivity rate (Table 1).
 
@@
 
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-#resulttable=readRDS("../../results/summarytable.rds")
-#knitr::kable(resulttable)
-```
+## Model analysis
 
-![Distribution of recorded seven-day moving average for each day](images/7DMA_histo.png)
-
-![Distribution of viral load per sample collection per target](images/total_copies_histo.png)
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-#knitr::include_graphics(here("results","height_weight_stratified.png"))
-```
-
-![Clinical case seven-day moving average over time](images/7DMA_time_plot.png)
-
-![Wastewater viral load over time](images/avg_total_copies_time.png)
-
-## Basic statistical analysis
-
-![Linear regression of viral load vs case seven-day moving average](images/total_copies_avg_vs_7DMA.png)
-
-## Full analysis
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-#resulttable2 = readRDS(here("results","resulttable2.rds"))
-#knitr::kable(resulttable2)
+```{r, echo=FALSE, fig.cap="Observed Clinical Cases vs Wastewater Predictions"}
+include_graphics(here("figures/modeling/assaypos_7dma.png"))
 ```

### 2023-04-26 code/analysis_code/exploratory/exploratory_analysis.qmd
update manuscript and readme

@@
   ggplot(aes(x = date, y = case_7dma)) +
-  geom_line(size =1.5, color = '#4c005c') +
+  stat_smooth(size =1.5, color = '#4c005c', span = .1) +
+  geom_point(color = '#4c005c') +
   xlab(label = "Date") +
@@
   ggplot(aes(date, pos_rate_tr)) +
-geom_line(size =1.5, color = '#634499') +
+stat_smooth(size =1.5, color = '#634499',span = .1 ) +
+  geom_point(color = "#634499") +
   xlab(label = "Date") +
@@
 data_clean %>% ggplot(aes(date, test_admin)) +
-  geom_line(size = 1.5) +
+  stat_smooth(size = 1.5, span = .1, color = "#2bcdef" ) +
+  geom_point(color = "#2bcdef") +
+    xlab(label = "Date") +
+  ylab(label = "Daily Tests Administered") +
   theme_light()
@@
   ggplot(aes(pos_rate_tr,percent_pos_test)) +
-  geom_point() +
-  stat_poly_line() +
+ geom_point(color = "#80256D") +
+  stat_poly_line(color = "#80256D") +
   stat_poly_eq() +

### 2023-04-26 code/modeling_code/2_log10VL_7DMA_model.Rmd
update manuscript and readme

@@
 test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
+test_dates <- readRDS(here("data/modeling_data/test_dates.RDS"))
 ```
@@
 
-## Create model using workflow
+## Fit model
 
@@
 ```{r}
-aug_test <- augment(fit_7dma, train_data)
-aug_test %>% select(case_7dma, .pred)
+aug_train <- augment(fit_7dma, train_data)
+aug_train %>% select(case_7dma, .pred)
 ```
 
-## Use RMSE and RSQ to evaluate model performance
+## Cross validation
 
 ```{r}
-rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
-rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
+folds <- vfold_cv(train_data, v = 5, repeats = 5, strata = case_7dma)
+
+cv_7dma <- fit_resamples(workflow_7dma, resamples = folds)
+collect_metrics(cv_7dma)
+```
+
+## Check null model
+
+```{r}
+workflow_null <- workflow() %>% 
+  add_model(null_mod) %>% 
+  add_recipe(recipe_7dma)
+
+fit_null_cv <- fit_resamples(workflow_null, resamples = folds)
+
+collect_metrics(fit_null_cv)
+```
+
+## Use RMSE and RSQ to evaluate model performance on training data
+
+```{r}
+rmse <- aug_train %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_train %>% rsq(truth = case_7dma, .pred)
 log10VL_7DMA_metrics <- full_join(rmse, rsq)
@@
 ```{r}
-aug_test %>% ggplot(aes(case_7dma, .pred)) +
+aug_train %>% ggplot(aes(case_7dma, .pred)) +
   geom_point() +
@@
 ```{r}
-aug_residual <- aug_test %>% mutate(residual = case_7dma - .pred)
+aug_train <- aug_test %>% mutate(residual = case_7dma - .pred)
 aug_residual %>% ggplot(aes(.pred, residual)) +
@@
 ```
+
+## Final fit on test data, then on test dates
+
+```{r}
+#testing data
+fit_test_data <- workflow_7dma %>% 
+  fit(data = test_data)
+
+aug_test <- augment(fit_test_data, test_data)
+aug_test %>% select(case_7dma, .pred)
+
+rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
+test_metrics <- full_join(rmse, rsq)
+
+test_metrics
+```
+
+```{r}
+#testing dates
+fit_test_dates <- workflow_7dma %>% 
+  fit(data = test_dates)
+
+aug_test_dates <- augment(fit_test_dates, test_dates)
+aug_test_dates %>% select(case_7dma, .pred)
+
+rmse <- aug_test_dates %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test_dates %>% rsq(truth = case_7dma, .pred)
+test_dates_metrics <- full_join(rmse, rsq)
+
+test_dates_metrics
+```
+
+## Plot testing dates predictions vs truth
+
+```{r}
+aug_test_dates %>% ggplot(aes(x=date)) +
+   stat_smooth(aes(y=case_7dma), method = "loess", span = 0.5, color = "black") +
+  stat_smooth(aes(y=.pred), method = "loess", span = 0.5)
+```

### 2023-04-26 code/modeling_code/4_AssayPos_7DMA_model.Rmd
update manuscript and readme

@@
 test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))
+full_data <- readRDS(here("data/processed_data/final_sum_combined_data.RDS"))
 ```
@@
 aug_test %>% select(case_7dma, .pred)
+```
+
+## Cross validation
+
+```{r}
+folds <- vfold_cv(train_data, v = 5, repeats = 5, strata = case_7dma)
+
+cv_7dma <- fit_resamples(workflow_7dma, resamples = folds)
+collect_metrics(cv_7dma)
+```
+
+## Check null model
+
+```{r}
+workflow_null <- workflow() %>% 
+  add_model(null_mod) %>% 
+  add_recipe(recipe_7dma)
+
+fit_null_cv <- fit_resamples(workflow_null, resamples = folds)
+
+collect_metrics(fit_null_cv)
 ```
@@
 ```
+
+```{r}
+#full data
+fit_full <- workflow_7dma %>% 
+  fit(data = full_data)
+
+aug_full <- augment(fit_full, full_data)
+aug_full %>% select(case_7dma, .pred)
+
+rmse <- aug_full %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_full %>% rsq(truth = case_7dma, .pred)
+test_full_metrics <- full_join(rmse, rsq)
+
+test_full_metrics
+```
+
+```{r}
+aug_full %>% ggplot(aes(x=date)) +
+ stat_smooth(aes(y=case_7dma), method = "loess", span = 0.1, color = "black", fill = "black") +
+  stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
+  theme_light()
+  ggsave(here("figures/modeling/assaypos_7dma."))
+```

### 2023-04-26 code/modeling_code/6_AssayPos_VL_7DMA_model.Rmd
update manuscript and readme

--- 2023-04-26 code/modeling_code/4_AssayPos_7DMA_model.Rmd
@@
 ---
-title: "assaypos_7dma_model"
+title: "assaypos_case_7dma_model"
 author: "Leah Lariscy"
@@
 
-Outcome: Clinical case seven day moving average
+Outcome: PCR test positivity rate (daily)
 
@@
 
-This is where we will set assaypos as predictor and 7dma as outcome
+This is where we will set assaypos as predictor and case_7dma as outcome
 
 ```{r}
-recipe_7dma <- recipe(case_7dma ~ pos_rate_tr, data = train_data)
+recipe_case_7dma <- recipe(case_7dma ~ pos_rate_tr + log10_VL, data = train_data)
 ```
@@
 ```{r}
-workflow_7dma <- workflow() %>% 
+workflow_case_7dma <- workflow() %>% 
   add_model(lm_mod) %>% 
-  add_recipe(recipe_7dma)
+  add_recipe(recipe_case_7dma)
 ```
@@
 set.seed(626)
-fit_7dma <- workflow_7dma %>% 
+fit_case_7dma <- workflow_case_7dma %>% 
   fit(data = train_data)
 
-tidy(fit_7dma)
+tidy(fit_case_7dma)
 ```
@@
 ```{r}
-aug_test <- augment(fit_7dma, train_data)
+aug_test <- augment(fit_case_7dma, train_data)
 aug_test %>% select(case_7dma, .pred)
-```
-
-## Cross validation
-
-```{r}
-folds <- vfold_cv(train_data, v = 5, repeats = 5, strata = case_7dma)
-
-cv_7dma <- fit_resamples(workflow_7dma, resamples = folds)
-collect_metrics(cv_7dma)
-```
-
-## Check null model
-
-```{r}
-workflow_null <- workflow() %>% 
-  add_model(null_mod) %>% 
-  add_recipe(recipe_7dma)
-
-fit_null_cv <- fit_resamples(workflow_null, resamples = folds)
-
-collect_metrics(fit_null_cv)
 ```
@@
 rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
-assaypos_7dma_metrics <- full_join(rmse, rsq)
+assaypos_case_7dma_metrics <- full_join(rmse, rsq)
 
-saveRDS(assaypos_7dma_metrics, here("data/modeling_data/metrics/assaypos_7dma_performance.RDS"))
+saveRDS(assaypos_case_7dma_metrics, here("data/modeling_data/metrics/assaypos_case_7dma_performance.RDS"))
 
-assaypos_7dma_metrics
+assaypos_case_7dma_metrics
 
@@
   stat_smooth(method = "lm")
-ggsave(here("data/modeling_data/metrics/assaypos_7dma_predict.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_case_7dma_predict.png"))
 ```
@@
 
-ggsave(here("data/modeling_data/metrics/assaypos_7dma_residual.png"))
+ggsave(here("data/modeling_data/metrics/assaypos_case_7dma_residual.png"))
 ```
@@
 #testing data
-fit_test_data <- workflow_7dma %>% 
+fit_test_data <- workflow_case_7dma %>% 
   fit(data = test_data)
@@
 #testing dates
-fit_test_dates <- workflow_7dma %>% 
+fit_test_dates <- workflow_case_7dma %>% 
   fit(data = test_dates)
@@
   stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
-  theme_light()
-  ggsave(here("figures/modeling/assaypos_7dma."))
+  theme_light() +
+
 ```

### 2023-04-26 code/modeling_code/dont_run/PPBoth_DailyPosTest_model.Rmd
update manuscript and readme

--- 2023-04-24 code/modeling_code/PPBoth_7DMA_model.Rmd
@@
 
-```{r, message=FALSE}
+```{r}
 library(tidyverse)
@@
 library(here)
-library(ggpmisc)
 ```
@@
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ percent_pos, data = train_data)
+recipe_7dma <- recipe(pcr_pos ~ percent_pos, data = train_data)
 ```
@@
 aug_test <- augment(fit_7dma, train_data)
-aug_test %>% select(pcr_pos_7dma, .pred)
+aug_test %>% select(pcr_pos, .pred)
 ```
@@
 ```{r}
-rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
-rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
-huber <- aug_test %>% huber_loss(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test %>% rmse(truth = pcr_pos, .pred)
+rsq <- aug_test %>% rsq(truth = pcr_pos, .pred)
+huber <- aug_test %>% huber_loss(truth = pcr_pos, .pred)
 PP_7DMA_metrics <- bind_rows(rmse, rsq, huber)
@@
 ```{r, message=FALSE}
-aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
+aug_test %>% ggplot(aes(pcr_pos, .pred)) +
   geom_point() +
-  stat_p
-  stat
+  stat_smooth(method = "lm")
 ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
@@
 
-## Plot residual values vs predicted values
+## Plot residual values
 
 ```{r, message=FALSE}
-aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
+aug_residual <- aug_test %>% mutate(residual = pcr_pos - .pred)
 aug_residual %>% ggplot(aes(.pred, residual)) +
-  geom_point() +
-  stat_poly_line()
+  geom_point()
 
 ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
-```
-
-## Plot residual values vs actual values
-
-```{r}
-aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
-aug_residual %>% ggplot(aes(pcr_pos_7dma, residual)) +
-  geom_point() +
-  stat_poly_line() +
-  stat_poly_eq() 
 ```
@@
 aug_test %>% ggplot(aes(x=date)) +
-  geom_point(aes(y=pcr_pos_7dma), color = "blue") +
-  geom_line(aes(y=pcr_pos_7dma), color = "blue") +
+  geom_point(aes(y=pcr_pos), color = "blue") +
+  geom_line(aes(y=pcr_pos), color = "blue") +
   geom_point(aes(y=.pred), color = "green") +
@@
 aug_test2 <- augment(fit_7dma, test_data)
-aug_test2 %>% select(pcr_pos_7dma, .pred)
+aug_test2 %>% select(pcr_pos, .pred)
 ```
@@
 aug_test2 %>% ggplot(aes(x=date)) +
-  geom_point(aes(y=pcr_pos_7dma), color = "blue") +
-  geom_line(aes(y=pcr_pos_7dma), color = "blue") +
+  geom_point(aes(y=pcr_pos), color = "blue") +
+  geom_line(aes(y=pcr_pos), color = "blue") +
  geom_point(aes(y=.pred), color = "green") +
@@
 ```{r}
-rmse <- aug_test2 %>% rmse(truth = pcr_pos_7dma, .pred)
-rsq <- aug_test2 %>% rsq(truth = pcr_pos_7dma, .pred)
-huber <- aug_test2 %>% huber_loss(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test2 %>% rmse(truth = pcr_pos, .pred)
+rsq <- aug_test2 %>% rsq(truth = pcr_pos, .pred)
+huber <- aug_test2 %>% huber_loss(truth = pcr_pos, .pred)
 PP_7DMA_test_metrics <- bind_rows(rmse, rsq, huber)
@@
 aug_test3 <- augment(fit_7dma, test_dates_both)
-aug_test3 %>% select(pcr_pos_7dma, .pred)
+aug_test3 %>% select(pcr_pos, .pred)
 ```
@@
 ```{r}
-rmse <- aug_test3 %>% rmse(truth = pcr_pos_7dma, .pred)
-rsq <- aug_test3 %>% rsq(truth = pcr_pos_7dma, .pred)
-huber <- aug_test3 %>% huber_loss(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test3 %>% rmse(truth = pcr_pos, .pred)
+rsq <- aug_test3 %>% rsq(truth = pcr_pos, .pred)
+huber <- aug_test3 %>% huber_loss(truth = pcr_pos, .pred)
 PP_7DMA_BA5_metrics <- bind_rows(rmse, rsq, huber)
@@
 aug_test3 %>% ggplot(aes(x=date)) +
-  geom_line(aes(y=pcr_pos_7dma), color = "#621940", size = 2) +
+  geom_line(aes(y=pcr_pos), color = "#621940", size = 2) +
   theme_light()
@@
  aug_test3 %>% ggplot(aes(x=date)) +
-   geom_line(aes(y=pcr_pos_7dma), color = "#1c246c", size = 2) +
+   geom_line(aes(y=pcr_pos), color = "#1c246c", size = 2) +
    geom_line(aes(y=.pred), color = "#ec4998", size = 2) +
    theme_light() +
-     ylab(label = "Predicted 7-Day Moving Average") +
+     ylab(label = "Predicted Daily Clinical Cases") +
   xlab(label = "Date") +
@@
 
-ggsave(here("figures/modeling/predicted_7DMA.png"))
+ggsave(here("figures/modeling/predicted_DailyCase.png"))
 ```

### 2023-04-26 code/modeling_code/dont_run/PPN1_7DMA_model.Rmd
update manuscript and readme

--- 2023-04-24 code/modeling_code/PPBoth_7DMA_model.Rmd
@@
 ```{r}
-train_data <- readRDS(here("data/modeling_data/training_data_both.RDS"))
-test_data <- readRDS(here("data/modeling_data/testing_data_both.RDS"))
+train_data <- readRDS(here("data/modeling_data/training_data_n1.RDS"))
+test_data <- readRDS(here("data/modeling_data/testing_data_n1.RDS"))
 ```
@@
 ```{r}
-set.seed(828)
+set.seed(626)
 fit_7dma <- workflow_7dma %>% 
@@
 
-```{r, message=FALSE}
+```{r}
 aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
   geom_point() +
-  stat_p
-  stat
+  stat_smooth(method = "lm")
 ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
@@
 
-## Plot residual values vs predicted values
-
-```{r, message=FALSE}
-aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
-aug_residual %>% ggplot(aes(.pred, residual)) +
-  geom_point() +
-  stat_poly_line()
-
-ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
-```
-
-## Plot residual values vs actual values
+## Plot residual values
 
@@
 aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
-aug_residual %>% ggplot(aes(pcr_pos_7dma, residual)) +
-  geom_point() +
-  stat_poly_line() +
-  stat_poly_eq() 
+aug_residual %>% ggplot(aes(.pred, residual)) +
+  geom_point()
+
+ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
 ```
@@
 ```{r}
-aug_test3 <- augment(fit_7dma, test_dates_both)
+aug_test3 <- augment(fit_7dma, test_dates_n1)
 aug_test3 %>% select(pcr_pos_7dma, .pred)
@@
 aug_test3 %>% ggplot(aes(x=date)) +
-  geom_line(aes(y=pcr_pos_7dma), color = "#621940", size = 2) +
-  theme_light()
-
+  geom_point(aes(y=pcr_pos_7dma), color = "blue") +
+  geom_line(aes(y=pcr_pos_7dma), color = "blue") +
+ geom_point(aes(y=.pred), color = "green") +
+  geom_line(aes(y=.pred), color = "green")
 ```
-
-```{r, message=FALSE}
- aug_test3 %>% ggplot(aes(x=date)) +
-   geom_line(aes(y=pcr_pos_7dma), color = "#1c246c", size = 2) +
-   geom_line(aes(y=.pred), color = "#ec4998", size = 2) +
-   theme_light() +
-     ylab(label = "Predicted 7-Day Moving Average") +
-  xlab(label = "Date") +
-  theme_light()
-
-ggsave(here("figures/modeling/predicted_7DMA.png"))
-```

### 2023-04-26 code/modeling_code/dont_run/PP_PosTest_model.Rmd
update manuscript and readme

--- 2023-03-20 code/modeling_code/PP_PosTest_model.Rmd
@@
 ```{r}
-train_data <- readRDS(here("data/modeling_data/training_data.RDS"))
+train_data <- readRDS(here("data/modeling_data/training_data.RDS")) %>% 
+  as.data.frame()
 test_data <- readRDS(here("data/modeling_data/testing_data.RDS"))

### 2023-04-26 code/modeling_code/dont_run/log10VL_PP_7DMA_model.Rmd
update manuscript and readme

--- 2023-03-20 code/modeling_code/log10VL_PP_7DMA_model.Rmd
@@
 ```{r}
-recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL + pos_rate_tr, data = train_data)
+recipe_7dma <- recipe(case_7dma ~ log10_VL + pos_rate_tr, data = train_data)
 ```
@@
 ```{r}
-rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
-rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
+rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
+rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
 log10VL_PP_7DMA_metrics <- full_join(rmse, rsq)

### 2023-04-26 products/manuscript/Manuscript.qmd
update manuscript and readme

@@
 format:
-  html:
+  docx:
     toc: false
@@
 
-Paragraph 1: Impact of COVID-19 on public health, clinical case under-reporting
+At the time of writing it has been three years since the COVID-19 Pandemic was declared, followed by a lockdown across the globe.
+To date, the World Health Organization reports over 760 million confirmed cases and 6.9 million deaths globally [@WHOCoronavirusCOVID19].
+The causative agent, severe acute respiratory syndrome coronavirus-2 (SARS-CoV-2), is a positive-sense single-stranded RNA zoonotic virus that infects cells via its spike proteins or "crowns" [@vkovskiCoronavirusBiologyReplication2021].
+When infected by this airborne pathogen, people report experiencing a wide range of mild to severe symptoms, including shortness of breath, cough, head and body aches, and loss of taste or smell.
+Older adults and others with underlying issues are at higher risk of experiencing more severe outcomes, including death [@COVID19YourHealth2020].
+However, long-term effects of COVID-19 on the body are not well understood yet and it may be years before enough research has been done to well-characterize these impacts.
+The effects of this pandemic have been far-reaching, with influence practically in every aspect of society, indicating a need to keep a close eye on this infectious disease.
+However, as the pandemic progresses, clinical reporting continues to dwindle, leaving researchers needing a better means of collecting data on the spread of the virus.
 
-Paragraph 2: Wastewater surveillance is possible because of excretory functions and stability of viral particles in wastewater, surveillance prevalence and correlations with clinical case data
+Early on in the pandemic, wastewater-based epidemiology (WBE) was identified as a promising method for detection of SARS-CoV-2 in a population.
+This is due to the ability of SARS-CoV-2 particles to shed from the body during and after infection through excretory functions, namely stool production [@wuProlongedPresenceSARSCoV22020].
+Viral particles in wastewater are able to be detected via Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) by amplification of universally-conserved gene targets on the SARS-CoV-2 genome [@ahmedFirstConfirmedDetection2020].
+The use of universal targets allows for the detection of the virus regardless of what variant is present.
+As of writing, over 4,000 sites worldwide have contributed to wastewater monitoring of SARS-CoV-2 [@COVIDPoops19].
+Not only is WBE an effective way of detecting viral presence, but quantification of wastewater viral load has been shown to correlate strongly with COVID-19 clinical cases when reporting efforts were strong, indicating the potential for it to be used as an alternative disease monitoring method.
 
@@
 Through exploration of time series and wastewater-clinical correlations, prominent relationships were identified for further analysis.
-Wastewater variables were used to train linear regression models to predict clinical observations.
+This also allowed for identification of an appropriate data split, whereby earlier observations were used to predict later ones.
+Wastewater variables were then used to train single-variate linear regression models to predict clinical observations.
+Multi-variate wastewater models were also created in the same manor to predict clinical observations.
 Model performance metrics (RSQ and RMSE) were assessed against each other and against a null model.
 Cross-validation was used to validate performances.
-The best model was then determined and used to predict values in a new set of data.
+The best model was then determined and used to predict values in data that was not used for training.
 
@@
 
+Comparison of wastewater and clinical variable time-series showed that these variables, although independent of one another, follow the same trends over time (Figures 1 & 2, Supplementary Figures).
+However, this changes in the time-series of administered clinical tests after the first few months of 2022 where there is nearly a complete drop-off in testing (Figure 3).
+This observation led to the decision to train linear regression models on data collected prior to February 28, 2022 and predict data collected after this date.
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 1. Statistics on clinical and wastewater variables of interest"}
+
+readRDS(here("results/manuscript_table1.rds"))
+
+```
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 1. Wastewater viral load time-series"}
+
+knitr::include_graphics(here("figures/exploratory/covariate/percent_pos_time_plot.png"))
+```
+
+```{r,  echo=FALSE, message=FALSE, fig.cap="Figure 2. Clinical case 7-day moving average time-series"}
+
+knitr::include_graphics(here("figures/exploratory/covariate/7DMA_time_plot.png"))
+```
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 3. Daily COVID-19 tests administered time-series"}
+knitr::include_graphics(here("figures/exploratory/covariate/test_admin_time.png"))
+```
+
 ## Correlation analysis
@@
 Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where stronger wastewater-clinical correlations were observed between these variables than in others.
-These include wastewater viral load, wastewater assay positivity rate, clinical case seven-day moving average, and test administration positivity rate (Table 1).
+These include wastewater viral load, wastewater assay positivity rate, clinical case seven-day moving average, and clinical test administration positivity rate (Table 1).
+Of these correlations, the strongest were observed when wastewater assay positivity was included as opposed to viral load (Figures 4 & 5, Supplementary Figures).
+Although this is true, linear regression models were still trained using both viral load and assay positivity.
 
-```{r, echo=FALSE, message=FALSE}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 4. Wastewater qPCR assay positivity vs clinical case 7-day moving average"}
 
-readRDS(here("results/manuscript_table1.rds"))
+knitr::include_graphics(here("figures/exploratory/covariate/assaypos_vs_7DMA.png"))
+```
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 5. Wastewater qPCR assay positivity vs clinical test positivity"}
+
+knitr::include_graphics(here("figures/exploratory/covariate/assaypos_testpos.png"))
 ```
@@
 
-```{r, echo=FALSE, fig.cap="Observed Clinical Cases vs Wastewater Predictions"}
-include_graphics(here("figures/modeling/assaypos_7dma.png"))
+```{r, echo=FALSE, fig.cap="Figure 6. Observed Clinical Cases vs Wastewater Predictions"}
+
+knitr::include_graphics(here("figures/modeling/assaypos_7dma.png"))
 ```

### 2023-04-26 products/manuscript/supplement/Supplementary_Material.qmd
update manuscript and readme

@@
 #| echo: FALSE
-#resulttable1 = readRDS(here("results","resulttable1.rds"))
-#knitr::kable(resulttable1)
+resulttable1 = readRDS(here("results","resulttable1.rds"))
+knitr::kable(resulttable1)
 ```
@@
 #| echo: FALSE
-#knitr::include_graphics(here("results","height_weight.png"))
+knitr::include_graphics(here("results","height_weight.png"))
 ```
@@
 
-Any additional discussion regarding the supplementary material/findings.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
 {{< pagebreak >}}

### 2023-04-27 review/Hemme_Project_Review.qmd
rendered review

---
title: Project Review Template 
author: Hayley Hemme
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Wastewater Surveillance for SARS-CoV-2 Data Analysis Project

Name of project author(s): Leah Lariscy

Name of project reviewer: Hayley Hemme

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments
Overall very clear and background was given. I think that your proposed third paragraph would be great to tie the background up.

### Summary assessment
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

Presented very clearly!

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

Data is clearly described. While the name structure of your variables is very clear, I think it would be useful for readers to see the variables names (as named in R) alongside the description of variable in this section. ex: clinical case seven-day moving average (`case_7dma`)

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

good job!

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

Methods seemed appropriate; predictor was lagged to account for viral shedding prior case reporting. Variables selected were reasonable.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Very nice!

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Missing.

### Summary assessment
* major parts of discussion missing or wrong

## Further comments

Hi Leah. I accidentally closed out of the file containing my feedback as I ran through your project, but I think that this covers everything I found.

The biggest issue I ran into was that your files did not run independently of each other. For the most part, this was only an issue when running code chucks for that contained `test dates`. I think that perhaps the best way to resolve this issue would to simply load this file into every data frame. However, it seems like your files might have been intended to run one after another. I would be sure to talk about this in you `readme` file if this was the intention. While running the files sequentially did resolve the issue of code chucks calling `test dates`, I did encounter an issue due this... in 6_AssayPos_TestPos_model.Rmd incorrectly calls on a workflow object from another model.

Additionally, I would suggest to using unique names for object/ vectors created by different models I think this is especially important when so much of your code is identical in structure across files. I think the most notable ones were the `aug_x`, and `rsme`, and `rsq`.

Also, I could not get the null models to run in any of the files.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
### Feedback and Comments

There were several files that were no longer needed, however these were clearly labeled.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Overall, very well documented.

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

As far as reproduciblity, I ran into issues with the files running not running indepedently of each other... (as described previously)

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

Alternatives were not discussed, however, methods and analysis seemed thorough enough to investigate the hypothesis.

### Summary assessment
* strong level of thorougness

### 2023-05-01 Project_Review_Template.qmd

# Overview

Title of project: Wastewater Surveillance for SARS-CoV-2 Data Analysis Project

Name of project author(s): Leah Lariscy

Name of project reviewer: Connor Ross

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Background and introduction as well as the data sources section were very well written and provided the reader adequate information to understand the project and the context of the existing work

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

Question(s)/hypotheses that the author posed was clear and related to the dataset.

### Summary assessment

* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data and the methods from which it was collected and used are well described and give a great understanding of the context and direction of the project. Further, it was clear what the overall variables of interests were.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The data was very well cleaned and formatted for a meaningful analysis. The comments provided in the processing code was extremely useful to understand the authors thought process behind the decisions that were made while cleaning.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The analysis methods were appropriate for the data and appear to have been performed properly without errors. Performance measures were provided in the analysis code in the built models and explained the results well.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

I really enjoyed the use of plots/figures that the author generated. It made their results extremely clear and provided information needed to make an informed determination as to the results of the analysis conducted. Plots were very clean and well formatted. A true example of an idea publishable figure for manuscript.

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Discussion and conclusion statements were partly missing from the final manuscript and the manuscipt qmd, however, the author did well covering aspects of these sections in other areas of the document. Completion of sections 5 through 5.3 is needed.

### Summary assessment

* minor parts wrong, missing or unclear

## Further comments

Awesome job on your analysis! I'm especially impressed with your plots that you generated! Very high quality!

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The project was very well organized. Everything was easy to find and the README.md file explained everything perfectly.

### Summary assessment

* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Yes, the commentary in your code/analysis documents were very clear and provided excellent explainations for the decisions you made.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Results are fully reproducible, however, there were some minor errors in code blocking a fully automated 'run-all'. This, however, did not impact the significant outputs or the final product. Please see errors and files below:

processing.qmd (line 512):
Error in table1(~log10_VL + pos_rate_tr + case_7dma + pcr_pos + percent_pos_test,  : 
  could not find function "table1"

2_log10VL_7DMA_model.Rmd (line 81):
Error in is_model_spec(spec) : object 'null_mod' not found

2_log10VL_7DMA_model.Rmd (line 115):
Error in mutate(., residual = case_7dma - .pred) : 
  object 'aug_test' not found

4_AssayPos_TestPos_model.Rmd (line 81):
Error in is_model_spec(spec) : object 'null_mod' not found

4_AssayPos_TestPos_model.Rmd (line 182):
Error in `ggsave()`:
! `filename` has no file extension and `device` is "NULL".
Backtrace:
 1. ggplot2::ggsave(here("figures/modeling/assaypos_7dma."))

6_AssayPos_VL_7DMA_model.Rmd (line 157):
Error: Incomplete expression: aug_full %>% ggplot(aes(x=date)) +
  stat_smooth(aes(y=case_7dma), method = "loess", span = 0.1, color = "black", fill = "black") +
  stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
  theme_light() +

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

I could definitely see were you were going with this and way you made  the decisions that you made. Don't forget to complete the discussion sections of your manuscript!

### Summary assessment

* decent level of thoroughness

## Further comments

I really enjoyed the use of plots/figures that the author generated. It made their results extremely clear and provided information needed to make an informed determination as to the results of the analysis conducted. Plots were very clean and well formatted. A true example of an idea publishable figure for manuscript.

### 2023-05-01 code/modeling_code/2_log10VL_7DMA_model.Rmd
update manuscript

@@
 cv_7dma <- fit_resamples(workflow_7dma, resamples = folds)
-collect_metrics(cv_7dma)
+cv_7dma_metrics <- collect_metrics(cv_7dma)
+
+saveRDS(cv_7dma_metrics, here("figures/modeling/cv_7dma_metrics.rds"))
 ```
@@
 ```{r}
-aug_train <- aug_test %>% mutate(residual = case_7dma - .pred)
+aug_residual <- aug_train %>% mutate(residual = case_7dma - .pred)
 aug_residual %>% ggplot(aes(.pred, residual)) +
@@
 test_dates_metrics
+
+saveRDS(test_dates_metrics, here("figures/modeling/cv_7dma_test_metrics.rds"))
 ```
@@
 aug_test_dates %>% ggplot(aes(x=date)) +
-   stat_smooth(aes(y=case_7dma), method = "loess", span = 0.5, color = "black") +
-  stat_smooth(aes(y=.pred), method = "loess", span = 0.5)
+   stat_smooth(aes(y=case_7dma), method = "loess", span = 0.2, color = "black") +
+  stat_smooth(aes(y=.pred), method = "loess", span = 0.2)
 ```

### 2023-05-01 products/manuscript/Manuscript.qmd
update manuscript

@@
 Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.
-Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.
+Subsequently, these viral particles can be detected in wastewater influent and used to estimate the abundance of COVID-19 in a community.
 Wastewater-based detection methods have been utilized across the globe as an independent and parallel indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends.
 WBE methods may also become a primary tool in understanding transmission dynamics as case under-reporting increases.
+Many surveillance groups have already shown wastewater viral load to be predictive of population clinical cases, so this study will evaluate the predictive power of alternative wastewater metrics.
+This study also aims to address the utility of SARS-CoV-2 wastewater surveillance to predict viral prevalence within a population once clinical reporting of COVID-19 declines significantly.
 
@@
 
-Paragraph 3: Surveillance program background. Aims, County-level parallel analysis of wastewater surveillance data and reported clinical cases
+Our lab
 
@@
 This study aims to conduct a parallel analysis between SARS-CoV-2 wastewater surveillance data and reported COVID-19 clinical case data in Athens-Clarke County, Georgia.
-It is expected that wastewater quantifications and reported case data will have a positive linear relationship.
-It is also hypothesized that certain wastewater-clinical relationships will be stronger than others, see Table 1 for variables of interest.
+Wastewater quantifications and reported case data are expected to have a positive linear relationship and it is hypothesized that certain wastewater-clinical relationships will be stronger than others, see Table 1 for variables of interest.
+It is also hypothesized that correlations will begin to decrease as the time series progresses due to diminishing clinical reporting.
 
@@
 Data sets for each standard curve was transformed separately before being combined.
-After combining all qPCR data, copies per liter (cp/L) of wastewater was calculated for each technical replicate using Equation 1.
+After combining all qPCR data, copies per liter (cp/L) of wastewater was calculated for each technical replicate (Equation 1).
 Average cp/L of wastewater was determined by taking the mean of all technical replicates per sample per gene target.
@@
 
-readRDS(here("results/manuscript_table1.rds"))
-
+table1 <- readRDS(here("results/manuscript_table1.rds"))
+knitr::kable(table1)
 ```
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure 1. Wastewater viral load time-series"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 1. Wastewater qPCR assay percent positivity time-series"}
 
@@
 
+Out of each feature combination, the model predicting seven-day moving average (7DMA) by wastewater assay percent positivity (APP) had the best performance metrics, confirmed by five-fold cross validation repeated five times (Table 2, Supplementary Tables).
+Metrics were also stable across a set of test data prior to February 28, 2022 that was withheld from model training.
+Thus, the 7DMA-APP model was selected to be fit to the test dates (after February 28, 2022).
+The metrics of this fit were also assessed, of which RMSE did not out-perform the null model (Table 3, Supplementary Table).
+
+```{r, echo=FALSE, fig.cap= "Table 2. 7DMA-APP Performance metrics when fit to training data"}
+cv_metrics <- readRDS(here("figures/modeling/cv_7dma_metrics.rds"))
+cv_metrics <- cv_metrics %>% mutate("Metric" = .metric, "Estimator" = .estimator, "Mean" = mean, "Standard Error" = std_err) %>% 
+  select(c(Metric, Estimator, Mean, n, "Standard Error"))
+knitr::kable(cv_metrics)
+```
+
+```{r, echo=FALSE, fig.cap= "Table 3. 7DMA-APP Performance metrics when fit to dates after February 28, 2022"}
+test_metrics <- readRDS(here("figures/modeling/cv_7dma_test_metrics.rds"))
+test_metrics <- test_metrics %>% mutate("Metric" = .metric, "Estimator" = .estimator, "Estimate" = .estimate) %>% 
+  select(c(Metric, Estimator, Estimate))
+knitr::kable(test_metrics)
+```
+
 ```{r, echo=FALSE, fig.cap="Figure 6. Observed Clinical Cases vs Wastewater Predictions"}
@@
 
-*Summarize what you did, what you found and what it means.*
+There were more significant correlations determined between certain variables than others, but SARS-CoV-2 wastewater surveillance metrics overall were shown to be predictive of clinical trends.
+These findings also suggest that qPCR assay positivity could be an alternative measure to viral load in relating wastewater data to population transmission levels of COVID-19.
+This is valuable because this metric involves simpler calculations and could be implemented in instances where treatment plant flow data is unavailable.
+It is also worth noting that wastewater-clinical correlations were higher in the beginning of the pandemic when testing efforts were more robust, whereas correlations were weaker in later dates.
+This, coupled with the finding that case reporting in Clarke County significantly decreased after February 2022, implicates this method of wastewater surveillance as a reliable means of understanding community disease risk when clinical data is not.
 
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
+The strength of this study lies in the ability of multiple wastewater metrics to predict clinical case data and that an alternative metric to viral load was identified.
+However, if improvements were to be made, more model tuning would be appropriate, seeing as certain model performance metrics did not far-excede the null model's performance.
+It is also worth noting that this could be in part due to the maxing out of assay percent positivity, making it difficult for the model to accurately estimate case abundance once the assay has become completely saturated.
 
@@
 
-*What are the main take-home messages?*
+Overall, the content of this work highlights the efficacy of these wastewater surveillance methods in capturing accurate trends in clinical cases within this county.
+Also, qPCR assay positivity was determined to be a slightly better predictor of clinical data than wastewater viral load.
+Subsequently, a linear regression model was trained on qPCR assay positivity to predict clinical seven-day moving average and was applied to a time-series of low case reporting.
+Due to upper-bound limitations of the model, the highest case levels were under-predicted both in the training and testing data.
+However, it is worth noting that during the Fall of 2022 when case reporting was low but assay positivity was high, the model estimated that case levels were much higher than what was reported.
 

### 2023-05-05 code/modeling_code/4_AssayPos_7DMA_model.Rmd
manuscript update

@@
   stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
-  theme_light()
-  ggsave(here("figures/modeling/assaypos_7dma."))
+  theme_light() +
+  xlab(label = "Date") +
+  ylab(label = "Clinical Case 7-Day Moving Average")
+
+  ggsave(here("figures/modeling/assaypos_7dma.png"))
 ```

### 2023-05-05 products/manuscript/Manuscript.qmd
manuscript update

@@
 To date, the World Health Organization reports over 760 million confirmed cases and 6.9 million deaths globally [@WHOCoronavirusCOVID19].
-The causative agent, severe acute respiratory syndrome coronavirus-2 (SARS-CoV-2), is a positive-sense single-stranded RNA zoonotic virus that infects cells via its spike proteins or "crowns" [@vkovskiCoronavirusBiologyReplication2021].
+The causative agent, Severe Acute Respiratory Syndrome Coronavirus-2 (SARS-CoV-2), is a positive-sense single-stranded RNA zoonotic virus that infects cells via its spike proteins or "crowns" [@vkovskiCoronavirusBiologyReplication2021].
 When infected by this airborne pathogen, people report experiencing a wide range of mild to severe symptoms, including shortness of breath, cough, head and body aches, and loss of taste or smell.
@@
 Not only is WBE an effective way of detecting viral presence, but quantification of wastewater viral load has been shown to correlate strongly with COVID-19 clinical cases when reporting efforts were strong, indicating the potential for it to be used as an alternative disease monitoring method.
-
-Our lab
 
@@
 
-{{< pagebreak >}}
-
 # Methods
@@
 As part of the qPCR panel, two gene targets on the N-gene of the SARS-CoV-2 genome were used for amplification.
-The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus \[\@\].
+The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus.
 Cycle threshold (Ct) values were then calculated for each PCR reaction, which were then used to estimate SARS-CoV-2 viral load (See Section 3.2 for more detail).
@@
 
-INSERT EQUATION 1
+$$
+C_W=C_R*(V_Q/V_D)*(V_T/V_R)*(V_E/V_W)*(1.0*10^6uL/1L)
+$$
 
-WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day.
+*Equation 1.
+Viral copies per liter of wastewater (C_W) was calculated by first multiplying copies per qPCR reaction (C_R) by volume of reaction (V_Q) divided by volume of converted DNA (V_D) in the reaction, then multiplying by volume of the reverse-transcription reaction (V_R) divided by volume of RNA extract (V_R) in that reaction, then multiplying by the elution volume of the RNA extraction (V_E) divided by the total volume of wastewater input (V_W).
+This yields copies per microliter of raw wastewater, which is finally converted to copies per liter.*
+
+WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day (Equation 2).
 qPCR data was then combined with WWTP data and total viral load per sample was calculated by multiplying average copies per liter by the flow rate.
@@
 County-level daily viral load was calculated by taking the sum of the average viral load from each of the three WWTP samples.
+
+$$
+VL_W=C_W*V_I
+$$
+
+*Equation 2. Wastewater viral load (VL_W) was determined by multiplying viral copies per liter of wastewater (C_W) by the total daily volume of influent flow in liters (V_I).*
 
@@
 
-{{< pagebreak >}}
-
 # Results
@@
 
-```{r,  echo=FALSE, message=FALSE, fig.cap="Figure 2. Clinical case 7-day moving average time-series"}
+```{r,  echo=FALSE, message=FALSE, fig.cap="Figure 2. Clinical case 7-day moving average time-series", fig.width=10, fig.height=8}
 
@@
 
-{{< pagebreak >}}
+In Figure 6, 7DMA predicted by the APP model is shown in blue, whereas the reported value is shown in black.
+The closest predictions appear to take place towards the beginning of the time-series, whereas the gap in predicted vs actual value widens as time progresses.
+It also appears that in many instances of high case abundance, the predicted values are under-estimated.
+This is likely because at a certain point, the qPCR assays reach nearly 100% and cannot accurately estimate values above that threshold.
+However, it is interesting that in the Fall of 2023 when clinical cases appeared to be severely under-reported (Figure 3), cases are estimated to be nearly as high as during the prior peak during the late summer months, whereas case abundance in the reported data is seemingly much lower.
 
@@
 The strength of this study lies in the ability of multiple wastewater metrics to predict clinical case data and that an alternative metric to viral load was identified.
-However, if improvements were to be made, more model tuning would be appropriate, seeing as certain model performance metrics did not far-excede the null model's performance.
-It is also worth noting that this could be in part due to the maxing out of assay percent positivity, making it difficult for the model to accurately estimate case abundance once the assay has become completely saturated.
+However, if improvements were to be made, more model tuning would be appropriate, seeing as certain model performance metrics did not far-exceed the null model's performance.
+It is also worth noting that this could be partially due to the maxing out of assay percent positivity, making it difficult for the model to accurately estimate case abundance once the assay has become completely saturated.
+Because of this, this particular model may be more useful at predicting periods of high vs low case abundance rather than trying to accurately estimate case counts.
+If done again, I would categorize each date in the wastewater time-series as either high, medium, or low viral abundance based on the distribution of the data, and build a categorical prediction model to predict periods of high, medium, or low case abundance.
 
@@
 Overall, the content of this work highlights the efficacy of these wastewater surveillance methods in capturing accurate trends in clinical cases within this county.
-Also, qPCR assay positivity was determined to be a slightly better predictor of clinical data than wastewater viral load.
+qPCR assay positivity was also determined to be a slightly better predictor of clinical data than wastewater viral load.
 Subsequently, a linear regression model was trained on qPCR assay positivity to predict clinical seven-day moving average and was applied to a time-series of low case reporting.
-Due to upper-bound limitations of the model, the highest case levels were under-predicted both in the training and testing data.
-However, it is worth noting that during the Fall of 2022 when case reporting was low but assay positivity was high, the model estimated that case levels were much higher than what was reported.
+Model performance metrics were better when the model was fit to the training data (prior to February 2022), but performance worsened significantly when fit to testing data (after February 2022).
+Due to upper-bound limitations though, the highest case levels were under-predicted both in the training and testing data.
+However, it is still worth noting that during the Fall of 2022 when case reporting was low but assay positivity was high, the model estimated that case levels were much higher than what was reported.
 

### 2023-05-05 review/review_Christian_Okitondo.qmd
manuscript update

---
title: Project Review
author: Christian Okitondo
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Wastewater Surveillance for SARS-CoV-2

Name of project author(s): Leah Laruscy

Name of project reviewer: Christian Okitondo

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The context of the project is well described in this document. The authors provide a comprehensive background on the COVID-19 pandemic, the causative agent, symptoms, and the impact of the pandemic on society. The authors also provide a detailed summary of previous and related work on wastewater-based detection methods that have been utilized across the globe as an independent and parallel indicator of SARS-CoV-2 viral prevalence. Proper referencing of existing work is also included.

The project is well placed in the context of existing work, and the authors clearly describe why the project was undertaken. The authors aim to conduct a parallel analysis between SARS-CoV-2 wastewater surveillance data and reported COVID-19 clinical case data in Athens-Clarke County, Georgia. The new information they hope to provide is a better understanding of the relationship between wastewater quantifications and reported case data, and to identify the specific wastewater-clinical relationships that are stronger than others. Overall, the context of the project is well-described and provides a clear understanding of the objectives and expected outcomes of the study.

### Summary assessment 
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments
The research questions are clearly described, and the hypotheses to be addressed are stated. The project aims to address the relationship between SARS-CoV-2 wastewater surveillance data and reported COVID-19 clinical case data in Athens-Clarke County, Georgia.

### Summary assessment
Question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data description is clear and the source is provided. The codebook is available in the readme.md file placed in the raw_data folder.

### Summary assessment

Source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments
The data cleaning/processing and exploration steps are well-explained and reasonable. The alternative methods are not discussed, but the author does discuss the limitations of the data and the possible sources of bias. The exploratory results are shown and meaningful. The paper provides a clear introduction, including background information about COVID-19, the rationale for the study, and the hypotheses to be tested. In the methods section, the author provides a detailed description of the data sources and acquisition process, as well as the statistical analyses used to test the hypotheses. Overall, the manuscript appears to be well-written and thoroughly researched.

### Summary assessment
Essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The analysis part is well-explained and presented in a good layout. Made good use of the machine learning lessons we learned.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments
The plots and tables for EDA are really nice. However, only the results from the correlation analysis was reported. Did not see results from the linear regression or any statistical tests mentioned in the methods section

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments
Discussion/Conclusion sections were not available.

### Summary assessment
* minor parts wrong, missing or unclear

## Further comments
Project heading in the right direction. With the results and discussion sections completed, it will make a strong one.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments
The project is well structured and labelled with appropriate names. There is no irrelevant or redundant file. It was fairly easy to reproduce everything by following the instructions.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Each step of the whole analysis is well-explained. There are plenty of comments justifying and stating the purpose of each line of code.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

All results are fully reproducible and the readme.me file under each folder provides clear guidance.

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Overall, the project is well writtend and properly analyzed. Once completed, it will make a good one.

### Summary assessment
* decent level of thoroughness
